---
title: "Group 2 Microbiome Analysis"
author: "Simar Jawanda"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "100%",
  dev = "png",
  fig.ext = "png",
  dpi = 300
)
```

# Step 1: Install and Load Packages

```{r}
required_packages <- c(
  "broom",
  "coin",
  "dplyr",
  "ggplot2",
  "patchwork",
  "rstatix",
  "tidyr",
  "tibble",
  "vegan",
  "scales",
  "ggh4x"
)

missing_packages <- required_packages[!required_packages %in% rownames(installed.packages())]

if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}

library(broom)
library(coin)
library(dplyr)
library(ggplot2)
library(rstatix)
library(tidyr)
library(tibble)
library(vegan)
library(scales)
library(ggh4x)

wrap_figure_title <- function(title_text, width = 56) {
  paste(strwrap(title_text, width = width), collapse = "\n")
}

finalize_thesis_plot <- function(plot, legend = c("none", "plain", "italic")) {
  legend <- match.arg(legend)

  if (!is.null(plot$labels$x) && identical(plot$labels$x, "Condition")) {
    plot <- plot + labs(x = NULL)
  }

  plot <- plot +
    labs(title = NULL) +
    theme(
      plot.title = element_blank(),
      plot.title.position = "panel",
      axis.line = element_line(linewidth = 1.1, colour = "black"),
      axis.ticks = element_line(linewidth = 0.95, colour = "black"),
      axis.ticks.length = grid::unit(0.18, "cm"),
      axis.title.x = element_text(size = 18, face = "bold", colour = "black"),
      axis.title.y = element_text(size = 18, face = "bold", colour = "black"),
      axis.text.x = element_text(size = 15.5, face = "bold", colour = "black"),
      axis.text.y = element_text(size = 15.5, face = "bold", colour = "black"),
      strip.background = element_rect(fill = "white", colour = "black", linewidth = 1.1),
      strip.text = element_text(size = 16.5, face = "bold", colour = "black")
    )

  if (legend != "none") {
    plot <- plot +
      guides(
        fill = guide_legend(ncol = 1, byrow = TRUE),
        colour = guide_legend(ncol = 1, byrow = TRUE),
        shape = guide_legend(ncol = 1, byrow = TRUE)
      ) +
      theme(
        legend.title = element_text(size = 20, face = "bold"),
        legend.text = element_text(size = 17, face = ifelse(legend == "italic", "italic", "plain")),
        legend.key.width = grid::unit(0.88, "cm"),
        legend.key.height = grid::unit(0.98, "cm"),
        legend.spacing.y = grid::unit(0.22, "cm"),
        legend.box = "vertical"
      )
  }

  plot
}

make_panel_label_map <- function(labels) {
  panel_labels <- paste0(LETTERS[seq_along(labels)], ". ", labels)
  stats::setNames(panel_labels, labels)
}

build_two_group_brackets <- function(data, facet_col, value_col, label_col) {
  data %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(facet_col))) %>%
    dplyr::summarise(
      bracket_y = max(.data[[value_col]], na.rm = TRUE) * 1.20 + 1e-06,
      label = dplyr::first(.data[[label_col]]),
      .groups = "drop"
    ) %>%
    dplyr::filter(!is.na(label), label != "") %>%
    dplyr::mutate(
      x_start = 1,
      x_end = 2,
      x_mid = 1.5,
      tick_drop = pmax(bracket_y * 0.04, 1e-06),
      label_y = bracket_y + pmax(bracket_y * 0.055, 1e-06)
    )
}

add_two_group_brackets <- function(plot, bracket_df, text_size = 6) {
  if (nrow(bracket_df) == 0) {
    return(plot)
  }

  plot +
    geom_segment(
      data = bracket_df,
      aes(x = x_start, xend = x_end, y = bracket_y, yend = bracket_y),
      inherit.aes = FALSE,
      linewidth = 0.9
    ) +
    geom_segment(
      data = bracket_df,
      aes(x = x_start, xend = x_start, y = bracket_y, yend = bracket_y - tick_drop),
      inherit.aes = FALSE,
      linewidth = 0.9
    ) +
    geom_segment(
      data = bracket_df,
      aes(x = x_end, xend = x_end, y = bracket_y, yend = bracket_y - tick_drop),
      inherit.aes = FALSE,
      linewidth = 0.9
    ) +
    geom_text(
      data = bracket_df,
      aes(x = x_mid, y = label_y, label = label),
      inherit.aes = FALSE,
      size = text_size,
      fontface = "bold"
    )
}

make_selected_genus_panel_plot <- function(boxplot_df, bracket_df, palette_values, panel_level) {
  panel_level <- as.character(panel_level)[1]

  panel_data <- boxplot_df %>%
    dplyr::filter(DisplayGenusPanel == panel_level)

  panel_brackets <- bracket_df %>%
    dplyr::filter(DisplayGenusPanel == panel_level)

  panel_label_data <- tibble::tibble(
    x_mid = 1.5,
    y_pos = Inf,
    label = panel_level
  )

  panel_plot <- ggplot(
    panel_data,
    aes(x = Condition, y = RelativeAbundance, fill = DisplayGenus)
  ) +
    stat_boxplot(
      geom = "errorbar",
      width = 0.22,
      linewidth = 1.0,
      coef = 1.5
    ) +
    geom_boxplot(
      outlier.shape = NA,
      width = 0.62,
      color = "black",
      linewidth = 1.0,
      coef = 1.5
    ) +
    geom_jitter(
      shape = 21,
      width = 0.07,
      size = 3.9,
      color = "black",
      stroke = 0.62,
      alpha = 0.82
    ) +
    scale_fill_manual(values = palette_values) +
    scale_x_discrete(
      limits = c("Control", "PS-NP"),
      expand = expansion(add = c(0.52, 0.52))
    ) +
    scale_y_continuous(expand = expansion(mult = c(0.02, 0.30))) +
    coord_cartesian(clip = "off") +
    theme_classic(base_size = 16) +
    labs(
      title = NULL,
      x = NULL,
      y = "Relative Abundance"
    ) +
    theme(
      axis.line = element_line(linewidth = 1.1, colour = "black"),
      axis.ticks = element_line(linewidth = 0.95, colour = "black"),
      axis.ticks.length = grid::unit(0.18, "cm"),
      axis.title.x = element_blank(),
      axis.title.y = element_text(size = 31.0, face = "bold", colour = "black", margin = margin(r = 10)),
      axis.text.x = element_text(size = 30.0, face = "bold", colour = "black"),
      axis.text.y = element_text(size = 28.0, face = "bold", colour = "black"),
      legend.position = "none",
      plot.margin = margin(t = 24, r = 18, b = 18, l = 12)
    )

  panel_plot <- add_two_group_brackets(panel_plot, panel_brackets, text_size = 7.8)

  panel_plot +
    geom_label(
      data = panel_label_data,
      aes(x = x_mid, y = y_pos, label = label),
      inherit.aes = FALSE,
      hjust = 0.5,
      vjust = 1.10,
      size = 8.8,
      fontface = "bold",
      label.size = 0.42,
      label.padding = grid::unit(0.14, "lines"),
      label.r = grid::unit(0.08, "lines"),
      fill = "white",
      colour = "black"
    )
}

make_selected_genus_panel_grid <- function(boxplot_df, bracket_df, palette_values, panel_levels, ncol = 2) {
  panel_levels <- as.character(panel_levels)

  plot_list <- lapply(panel_levels, function(panel_level) {
    make_selected_genus_panel_plot(boxplot_df, bracket_df, palette_values, panel_level)
  })

  while (length(plot_list) %% ncol != 0) {
    plot_list <- c(plot_list, list(patchwork::plot_spacer()))
  }

  patchwork::wrap_plots(plot_list, ncol = ncol, byrow = TRUE)
}


deterministic_extra_colors <- function(labels, chroma = 55, luminance = 70) {
  colors <- vapply(labels, function(lbl) {
    ints <- utf8ToInt(enc2utf8(lbl))
    hue <- sum(ints * seq_along(ints)) %% 360
    grDevices::hcl(h = hue, c = chroma, l = luminance)
  }, character(1))
  stats::setNames(colors, labels)
}

extract_taxonomy_rank <- function(x, rank_letter) {
  pattern <- paste0("(?:^|;)", rank_letter, "_{1,2}([^;]+)")
  value <- stringr::str_match(x, pattern)[, 2]
  value <- ifelse(is.na(value), "", value)
  trimws(gsub("_", " ", value))
}

build_parent_rank_label <- function(family, order, class_name, phylum) {
  invalid_or_generic <- function(v) {
    v == "" | is.na(v) | grepl("^[a-zA-Z]__$", v) |
      tolower(v) %in% c("unassigned", "unclassified", "unknown", "uncultured", "incertae sedis")
  }

  family_invalid <- invalid_or_generic(family)
  order_invalid <- invalid_or_generic(order)
  class_invalid <- invalid_or_generic(class_name)
  phylum_invalid <- invalid_or_generic(phylum)

  parent_taxon <- ifelse(
    !family_invalid, family,
    ifelse(
      !order_invalid, order,
      ifelse(!class_invalid, class_name, ifelse(!phylum_invalid, phylum, NA_character_))
    )
  )

  parent_rank <- ifelse(
    !family_invalid, "family",
    ifelse(
      !order_invalid, "order",
      ifelse(!class_invalid, "class", ifelse(!phylum_invalid, "phylum", NA_character_))
    )
  )

  ifelse(
    !is.na(parent_taxon) & !is.na(parent_rank),
    paste0(parent_rank, ": ", parent_taxon),
    NA_character_
  )
}

clean_taxon_label_composition <- function(x) {
  genus <- extract_taxonomy_rank(x, "g")
  family <- extract_taxonomy_rank(x, "f")
  order <- extract_taxonomy_rank(x, "o")
  class_name <- extract_taxonomy_rank(x, "c")
  phylum <- extract_taxonomy_rank(x, "p")

  genus_invalid <- genus == "" | is.na(genus) | grepl("^[a-z]__$", genus)
  parent_label <- build_parent_rank_label(family, order, class_name, phylum)
  genus_lower <- tolower(genus)
  raw_lower <- tolower(x)

  fallback_unresolved_label <- ifelse(
    grepl("unassigned", raw_lower),
    "Unassigned",
    ifelse(
      grepl("unclassified", raw_lower),
      "Unclassified",
      ifelse(
        grepl("uncultured", raw_lower),
        "Uncultured",
        ifelse(grepl("incertae[_ ]sedis", raw_lower), "Incertae Sedis", "Unclassified")
      )
    )
  )

  ifelse(
    !genus_invalid & !genus_lower %in% c("uncultured", "unassigned", "unclassified", "unknown", "incertae sedis"),
    genus,
    ifelse(
      genus_lower == "uncultured",
      ifelse(!is.na(parent_label), paste0("Uncultured (", parent_label, ")"), "Uncultured (taxonomy unavailable)"),
      ifelse(
        genus_lower %in% c("unassigned", "unclassified", "unknown", "incertae sedis"),
        ifelse(
          !is.na(parent_label),
          paste0(tools::toTitleCase(genus), " (", parent_label, ")"),
          paste0(tools::toTitleCase(genus), " (taxonomy unavailable)")
        ),
        ifelse(
          !is.na(parent_label),
          paste0("Unclassified (", parent_label, ")"),
          paste0(fallback_unresolved_label, " (taxonomy unavailable)")
        )
      )
    )
  )
}

build_shared_genus_palette_composition <- function(genus_names) {
  genus_palette <- c(
    "Lactobacillus" = "#8ECFC9",
    "Muribaculaceae" = "#F3C78D",
    "Streptococcus" = "#A8B6D9",
    "Turicibacter" = "#E8A0A6",
    "Faecalibaculum" = "#A9CFA5",
    "Odoribacter" = "#DCC7A1",
    "Bifidobacterium" = "#C8B6D8",
    "Rikenellaceae RC9 gut group" = "#E7B8D6",
    "Alistipes" = "#90B4C6",
    "Helicobacter" = "#D8A48F",
    "Akkermansia" = "#B7A0D8",
    "Bacteroides" = "#8FBE6D",
    "Clostridia UCG-014" = "#B48A8A",
    "Clostridium sensu stricto 1" = "#BFB6B0",
    "Lachnospiraceae NK4A136 group" = "#5B84B1",
    "Muribaculum" = "#9E9D68",
    "Candidatus Saccharimonas" = "#6F8FB4",
    "Desulfovibrio" = "#A77C95",
    "Dubosiella" = "#8AA24A",
    "Uncultured (Lachnospiraceae)" = "#9FB6C8",
    "Uncultured (Ruminococcaceae)" = "#AFC1D8",
    "Uncultured (Oscillospiraceae)" = "#8AAFD0",
    "Incertae Sedis (Ruminococcaceae)" = "#5FAE8D",
    "Uncultured (Bacillaceae)" = "#A6C0D8",
    "Uncultured (Christensenellaceae)" = "#B5C6D9",
    "Uncultured (Desulfovibrionaceae)" = "#9CAFC4",
    "Uncultured (Erysipelotrichaceae)" = "#B7C0CC",
    "Uncultured (Peptococcaceae)" = "#A9B4C7",
    "Unclassified (Lachnospiraceae)" = "#D5B3C6",
    "Unclassified (Clostridia)" = "#CFA4BF",
    "Unclassified (Erysipelatoclostridiaceae)" = "#DAB7C5",
    "Unclassified (taxonomy unavailable)" = "#A9C8E8",
    "Unassigned (taxonomy unavailable)" = "#C892A6",
    "Uncultured (taxonomy unavailable)" = "#C7CEDB",
    "Uncultured" = "#C7CEDB",
    "Unassigned" = "#C892A6",
    "Unassigned (Lachnospiraceae)" = "#E0B180",
    "Unclassified" = "#A9C8E8",
    "Other" = "#A6A6A6"
  )

  missing_genera <- setdiff(genus_names, names(genus_palette))

  if (length(missing_genera) > 0) {
    extra_cols <- deterministic_extra_colors(missing_genera)
    genus_palette <- c(genus_palette, extra_cols)
  }

  genus_palette
}

run_shapiro_by_condition <- function(data, value_col) {
  bind_rows(
    lapply(
      split(data[[value_col]], data$group_simple),
      function(values) {
        if (length(values) >= 3 && length(values) <= 5000) {
          test <- shapiro.test(values)
          tibble(
            n = length(values),
            statistic = unname(test$statistic),
            p_value = unname(test$p.value),
            method = unname(test$method)
          )
        } else {
          tibble(
            n = length(values),
            statistic = NA_real_,
            p_value = NA_real_,
            method = "Shapiro-Wilk not run; sample size outside 3-5000"
          )
        }
      }
    ),
    .id = "group_simple"
  ) %>%
    mutate(metric = value_col) %>%
    relocate(metric, .before = group_simple)
}

run_genus_pca <- function(genus_merged_df, metadata_cols, top_n = 50) {
  genus_abundance_only <- genus_merged_df %>%
    select(-all_of(metadata_cols)) %>%
    mutate(across(everything(), as.numeric))

  keep_genus <- vapply(
    genus_abundance_only,
    function(values) {
      sum(!is.na(values)) > 1 && sd(values, na.rm = TRUE) > 0
    },
    logical(1)
  )

  genus_abundance_filtered <- genus_abundance_only[, keep_genus, drop = FALSE]

  pca_obj <- prcomp(genus_abundance_filtered, center = TRUE, scale. = TRUE)

  variance_explained <- (pca_obj$sdev^2) / sum(pca_obj$sdev^2)
  variance_df <- tibble(
    PC = paste0("PC", seq_along(variance_explained)),
    variance_explained = variance_explained,
    cumulative_variance = cumsum(variance_explained)
  )

  loading_pc_count <- min(5, ncol(pca_obj$rotation))
  loading_weights <- variance_explained[seq_len(loading_pc_count)]
  loadings_matrix <- pca_obj$rotation[, seq_len(loading_pc_count), drop = FALSE]
  weighted_loading <- rowSums(abs(sweep(loadings_matrix, 2, loading_weights, `*`)))

  loadings_df <- as.data.frame(loadings_matrix) %>%
    rownames_to_column("Genus") %>%
    mutate(weighted_loading = weighted_loading[Genus]) %>%
    arrange(desc(weighted_loading))

  top_genera_df <- loadings_df %>%
    slice_head(n = min(top_n, nrow(loadings_df)))

  filtered_table <- genus_merged_df %>%
    select(all_of(metadata_cols), all_of(top_genera_df$Genus))

  list(
    variance_df = variance_df,
    loadings_df = loadings_df,
    top_genera_df = top_genera_df,
    filtered_table = filtered_table
  )
}

run_stratified_wilcoxon <- function(data, value_col) {
  test_data <- data %>%
    transmute(
      value = as.numeric(.data[[value_col]]),
      group_simple = factor(group_simple, levels = c("control", "psnp")),
      accession_number = factor(accession_number, levels = c("PRJNA944557", "PRJNA1357312"))
    ) %>%
    filter(
      !is.na(value),
      !is.na(group_simple),
      !is.na(accession_number)
    )

  test_result <- coin::wilcox_test(
    value ~ group_simple | accession_number,
    data = test_data,
    distribution = "asymptotic"
  )

  tibble(
    metric = value_col,
    statistic = as.numeric(statistic(test_result, type = "standardized")),
    p_value = as.numeric(coin::pvalue(test_result)),
    method = "Stratified Wilcoxon rank-sum test by accession_number"
  )
}
```

# Step 2: Set File Paths

```{r}
file_path <- "/Users/simarjawanda/Library/CloudStorage/OneDrive-UBC/QIIME2 DATA/Group 2/metadata_merged_group2_filtered_R_analysis.csv"
sample_id_col <- "Run"
condition_col <- "group_simple"
batch_col <- "accession_number"
size_col <- "psnp_size"
output_dir <- "/Users/simarjawanda/Library/CloudStorage/OneDrive-UBC/R Analysis/Group 2 Outputs"
top50_output_dir <- file.path(output_dir, "pca_top50")

if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

if (!dir.exists(top50_output_dir)) {
  dir.create(top50_output_dir, recursive = TRUE)
}
```

# Step 3: Read Filtered Metadata

```{r}
metadata_file <- "/Users/simarjawanda/Library/CloudStorage/OneDrive-UBC/QIIME2 DATA/Group 2/metadata_merged_group2_filtered_R_analysis.csv"
metadata_df <- read.csv(metadata_file, check.names = FALSE)

head(metadata_df[, c("Run", "group_simple", "psnp_size", "accession_number")])
table(metadata_df$group_simple, useNA = "ifany")
table(metadata_df$accession_number, useNA = "ifany")
table(metadata_df$group_simple, metadata_df$accession_number, useNA = "ifany")
```

# Step 4: Keep Only Control and PS-NP Samples

```{r}
metadata_df <- metadata_df %>%
  filter(group_simple %in% c("control", "psnp"))

head(metadata_df[, c("Run", "group_simple", "psnp_size", "accession_number")])
table(metadata_df$group_simple, useNA = "ifany")
table(metadata_df$accession_number, useNA = "ifany")
table(metadata_df$group_simple, metadata_df$accession_number, useNA = "ifany")
```

# Step 5: Read ASV Table

```{r}
asv_file <- "/Users/simarjawanda/Library/CloudStorage/OneDrive-UBC/QIIME2 DATA/Group 2/exported-asv-table/asv-table.tsv"

asv_df <- read.delim(
  asv_file,
  check.names = FALSE,
  skip = 1
)

dim(asv_df)
head(asv_df[, 1:6])
colnames(asv_df)[1:6]
```

# Step 6: Transpose ASV Table

```{r}
asv_matrix <- asv_df
rownames(asv_matrix) <- asv_matrix[, 1]
asv_matrix <- asv_matrix[, -1]

asv_matrix_t <- as.data.frame(t(asv_matrix))
asv_matrix_t$Run <- rownames(asv_matrix_t)
rownames(asv_matrix_t) <- NULL

dim(asv_matrix_t)
head(asv_matrix_t[, 1:6])
```

# Step 7: Merge Metadata and ASVs

```{r}
merged_df <- metadata_df %>%
  inner_join(asv_matrix_t, by = "Run")

dim(merged_df)
head(merged_df[, c("Run", "group_simple", "psnp_size", "accession_number")])

table(merged_df$group_simple, useNA = "ifany")
table(merged_df$accession_number, useNA = "ifany")
table(merged_df$group_simple, merged_df$accession_number, useNA = "ifany")
```

# Step 8: Keep Analysis Columns

```{r}
analysis_metadata <- merged_df %>%
  select(Run, group_simple, psnp_size, accession_number)

analysis_metadata$group_simple <- factor(analysis_metadata$group_simple, levels = c("control", "psnp"))
analysis_metadata$accession_number <- factor(
  analysis_metadata$accession_number,
  levels = c("PRJNA944557", "PRJNA1357312")
)

abundance_df <- merged_df %>%
  select(-all_of(colnames(metadata_df)))

dim(analysis_metadata)
dim(abundance_df)
```

# Step 9: Make Group Summary Table

```{r}
group_summary_df <- analysis_metadata %>%
  dplyr::count(accession_number, group_simple, name = "n")

group_summary_df
table(analysis_metadata$group_simple, analysis_metadata$accession_number)
```

# Step 10: Clean and Normalize Data

```{r}
abundance_df <- abundance_df %>%
  mutate(across(everything(), as.numeric))

abundance_df[is.na(abundance_df)] <- 0

min_value <- min(as.matrix(abundance_df)[as.matrix(abundance_df) > 0], na.rm = TRUE)
pseudocount <- min_value / 5

abundance_df <- abundance_df %>%
  mutate(across(everything(), ~ ifelse(. == 0, pseudocount, .)))

row_totals <- rowSums(abundance_df)
norm_abundance_df <- abundance_df / row_totals

summary(row_totals)
range(rowSums(norm_abundance_df))
pseudocount
```

# Step 11: Calculate Alpha Diversity

```{r}
shannon_index <- diversity(norm_abundance_df, index = "shannon")
simpson_index <- diversity(norm_abundance_df, index = "simpson")

alpha_df <- data.frame(
  Run = analysis_metadata$Run,
  Shannon = shannon_index,
  Simpson = simpson_index
)

alpha_meta_df <- alpha_df %>%
  left_join(analysis_metadata, by = "Run") %>%
  mutate(
    Condition = ifelse(group_simple == "psnp", "PS-NP", "Control"),
    Condition = factor(Condition, levels = c("Control", "PS-NP"))
  )

head(alpha_meta_df)
summary(alpha_meta_df$Shannon)
summary(alpha_meta_df$Simpson)
```

# Step 12: Make Shannon Boxplot

```{r shannon_plot_group2, fig.width=8.4, fig.height=6.8}
alpha_meta_df$Condition <- factor(alpha_meta_df$Condition, levels = c("Control", "PS-NP"))
shannon_p_value <- wilcox.test(Shannon ~ group_simple, data = alpha_meta_df, exact = FALSE)$p.value
shannon_significance_label <- ifelse(shannon_p_value < 0.05, "*", "")
shannon_annotation_y <- max(alpha_meta_df$Shannon, na.rm = TRUE) * 1.08

shannon_plot <- ggplot(alpha_meta_df, aes(x = Condition, y = Shannon, fill = Condition)) +
  stat_boxplot(
    geom = "errorbar",
    width = 0.18,
    linewidth = 0.8,
    coef = 1.5
  ) +
  geom_boxplot(
    outlier.shape = NA,
    width = 0.5,
    color = "black",
    linewidth = 0.8,
    coef = 1.5
  ) +
  geom_jitter(
    shape = 21,
    width = 0.08,
    size = 3.0,
    color = "black",
    stroke = 0.5,
    alpha = 0.8
  ) +
  scale_fill_manual(values = c("Control" = "#B7D9F2", "PS-NP" = "#F2B6A7")) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.14))) +
  annotate("text", x = 1.5, y = shannon_annotation_y, label = shannon_significance_label, size = 7, fontface = "bold") +
  theme_classic(base_size = 15, base_family = "Helvetica") +
  labs(
    title = wrap_figure_title("Shannon Alpha Diversity in Control and PS-NP Samples for Group 2"),
    x = "Condition",
    y = "Shannon Index"
  ) +
  theme(
    plot.title = element_text(size = 17, hjust = 0.5, face = "plain", lineheight = 1.05, margin = margin(b = 10)),
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 13, color = "black"),
    legend.position = "none"
  )

shannon_plot <- finalize_thesis_plot(shannon_plot, legend = "none")

shannon_plot
```

# Step 13: Make Simpson Boxplot

```{r simpson_plot_group2, fig.width=8.4, fig.height=6.8}
simpson_p_value <- wilcox.test(Simpson ~ group_simple, data = alpha_meta_df, exact = FALSE)$p.value
simpson_significance_label <- ifelse(simpson_p_value < 0.05, "*", "")
simpson_annotation_y <- max(alpha_meta_df$Simpson, na.rm = TRUE) * 1.08

simpson_plot <- ggplot(alpha_meta_df, aes(x = Condition, y = Simpson, fill = Condition)) +
  stat_boxplot(
    geom = "errorbar",
    width = 0.18,
    linewidth = 0.8,
    coef = 1.5
  ) +
  geom_boxplot(
    outlier.shape = NA,
    width = 0.5,
    color = "black",
    linewidth = 0.8,
    coef = 1.5
  ) +
  geom_jitter(
    shape = 21,
    width = 0.08,
    size = 3.0,
    color = "black",
    stroke = 0.5,
    alpha = 0.8
  ) +
  scale_fill_manual(values = c("Control" = "#B7D9F2", "PS-NP" = "#F2B6A7")) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.14))) +
  annotate("text", x = 1.5, y = simpson_annotation_y, label = simpson_significance_label, size = 7, fontface = "bold") +
  theme_classic(base_size = 15, base_family = "Helvetica") +
  labs(
    title = wrap_figure_title("Simpson Alpha Diversity in Control and PS-NP Samples for Group 2"),
    x = "Condition",
    y = "Simpson Index"
  ) +
  theme(
    plot.title = element_text(size = 17, hjust = 0.5, face = "plain", lineheight = 1.05, margin = margin(b = 10)),
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 13, color = "black"),
    legend.position = "none"
  )

simpson_plot <- finalize_thesis_plot(simpson_plot, legend = "none")

simpson_plot
```

# Step 14: Run Alpha Diversity Normality Checks

```{r}
shapiro_shannon <- run_shapiro_by_condition(alpha_meta_df, "Shannon")
shapiro_simpson <- run_shapiro_by_condition(alpha_meta_df, "Simpson")

mann_whitney_shannon <- wilcox.test(Shannon ~ group_simple, data = alpha_meta_df, exact = FALSE)
mann_whitney_simpson <- wilcox.test(Simpson ~ group_simple, data = alpha_meta_df, exact = FALSE)

stratified_wilcoxon_shannon <- run_stratified_wilcoxon(alpha_meta_df, "Shannon")
stratified_wilcoxon_simpson <- run_stratified_wilcoxon(alpha_meta_df, "Simpson")

shapiro_shannon
shapiro_simpson
mann_whitney_shannon
mann_whitney_simpson
stratified_wilcoxon_shannon
stratified_wilcoxon_simpson
```

# Step 15: Read Genus Relative Abundance Table

```{r}
genus_file <- "/Users/simarjawanda/Library/CloudStorage/OneDrive-UBC/QIIME2 DATA/Group 2/exported-genus-relfreq/genus-relfreq.tsv"

genus_df <- read.delim(genus_file, check.names = FALSE, skip = 1)

dim(genus_df)
head(genus_df[, 1:6])
colnames(genus_df)[1:6]
```

# Step 16: Transpose Genus Table and Merge with Metadata

```{r}
genus_matrix <- genus_df
rownames(genus_matrix) <- genus_matrix[, 1]
genus_matrix <- genus_matrix[, -1]

genus_matrix_t <- as.data.frame(t(genus_matrix))
genus_matrix_t$Run <- rownames(genus_matrix_t)
rownames(genus_matrix_t) <- NULL

genus_merged_df <- analysis_metadata %>%
  inner_join(genus_matrix_t, by = "Run")

dim(genus_merged_df)
```

# Step 17: Run PCA on the Genus Table and Select the Top 50 Genera

```{r}
group2_pca <- run_genus_pca(
  genus_merged_df = genus_merged_df,
  metadata_cols = c("Run", "group_simple", "psnp_size", "accession_number"),
  top_n = 50
)

pca_variance_df <- group2_pca$variance_df
pca_genus_loadings_df <- group2_pca$loadings_df
pca_top50_genera_df <- group2_pca$top_genera_df
genus_filtered_top50_df <- group2_pca$filtered_table

head(pca_variance_df, 10)
head(pca_top50_genera_df, 20)
```

# Step 18: Save Current Shapiro and PCA Filtering Outputs

```{r}
write.csv(metadata_df, file.path(output_dir, "group2_metadata.csv"), row.names = FALSE)
write.csv(analysis_metadata, file.path(output_dir, "group2_analysis_metadata.csv"), row.names = FALSE)
write.csv(group_summary_df, file.path(output_dir, "group2_group_summary.csv"), row.names = FALSE)
write.csv(abundance_df, file.path(output_dir, "group2_abundance_cleaned.csv"), row.names = FALSE)
write.csv(as.data.frame(norm_abundance_df), file.path(output_dir, "group2_abundance_normalized.csv"), row.names = TRUE)
write.csv(alpha_df, file.path(output_dir, "group2_alpha_diversity.csv"), row.names = FALSE)
write.csv(alpha_meta_df, file.path(output_dir, "group2_alpha_diversity_with_metadata.csv"), row.names = FALSE)
write.csv(shapiro_shannon, file.path(output_dir, "group2_shapiro_shannon_by_condition.csv"), row.names = FALSE)
write.csv(shapiro_simpson, file.path(output_dir, "group2_shapiro_simpson_by_condition.csv"), row.names = FALSE)
write.csv(pca_variance_df, file.path(top50_output_dir, "group2_pca_variance_explained.csv"), row.names = FALSE)
write.csv(pca_genus_loadings_df, file.path(top50_output_dir, "group2_pca_genus_loadings.csv"), row.names = FALSE)
write.csv(pca_top50_genera_df, file.path(top50_output_dir, "group2_pca_top50_genera.csv"), row.names = FALSE)
write.csv(genus_filtered_top50_df, file.path(top50_output_dir, "group2_genus_filtered_top50_for_downstream.csv"), row.names = FALSE)

ggsave(file.path(output_dir, "group2_shannon_boxplot.png"), plot = shannon_plot, width = 6.5, height = 5.5, dpi = 600)
ggsave(file.path(output_dir, "group2_simpson_boxplot.png"), plot = simpson_plot, width = 6.5, height = 5.5, dpi = 600)
```

# Step 19: Calculate Bray-Curtis Distances

```{r}
rownames(norm_abundance_df) <- analysis_metadata$Run

bray_dist <- vegdist(norm_abundance_df, method = "bray")

as.matrix(bray_dist)[1:5, 1:5]
```

# Step 20: Run PCoA

```{r}
pcoa_obj <- cmdscale(bray_dist, k = 2, eig = TRUE, add = TRUE)

pcoa_df <- as.data.frame(pcoa_obj$points)
colnames(pcoa_df) <- c("PCoA1", "PCoA2")
pcoa_df$Run <- rownames(pcoa_obj$points)

pcoa_plot_df <- pcoa_df %>%
  left_join(analysis_metadata, by = "Run") %>%
  mutate(
    Condition = ifelse(group_simple == "psnp", "PS-NP", "Control"),
    Condition = factor(Condition, levels = c("Control", "PS-NP"))
  )

eig <- pcoa_obj$eig
positive_eig <- eig[eig > 0]

pcoa1_percent <- round(100 * positive_eig[1] / sum(positive_eig), 1)
pcoa2_percent <- round(100 * positive_eig[2] / sum(positive_eig), 1)

head(pcoa_plot_df)
pcoa1_percent
pcoa2_percent
```

# Step 21: Make Batch-Aware PCoA Plot

```{r pcoa_plot_group2, fig.width=8.5, fig.height=6.5}
shape_values <- c(
  "PRJNA944557" = 24,
  "PRJNA1357312" = 22
)

pcoa_plot <- ggplot(pcoa_plot_df, aes(x = PCoA1, y = PCoA2)) +
  geom_point(
    aes(fill = Condition, shape = accession_number),
    size = 4.8,
    color = "black",
    stroke = 0.9,
    alpha = 0.95,
    position = position_jitter(width = 0.009, height = 0.009)
  ) +
  scale_fill_manual(
    values = c("Control" = "#B7D9F2", "PS-NP" = "#F2B6A7"),
    name = "Condition"
  ) +
  scale_shape_manual(
    values = shape_values,
    name = "Study"
  ) +
  guides(
    fill = guide_legend(
      order = 2,
      override.aes = list(
        shape = 21,
        size = 6.4,
        color = "black",
        stroke = 0.9,
        fill = c("#B7D9F2", "#F2B6A7")
      )
    ),
    shape = guide_legend(
      order = 1,
      override.aes = list(fill = "white", size = 6.8, color = "black", stroke = 0.9)
    )
  ) +
  theme_classic(base_size = 18) +
  labs(
    title = "Bray-Curtis PCoA of Group 2 Samples by Condition and Study Accession",
    x = paste0("PCoA1 (", pcoa1_percent, "%)"),
    y = paste0("PCoA2 (", pcoa2_percent, "%)")
  ) +
  theme(
    plot.title = element_text(size = 20, face = "plain", hjust = 0.5, lineheight = 1.05, margin = margin(b = 10)),
    legend.position = "right",
    plot.title.position = "plot"
  )

pcoa_plot <- finalize_thesis_plot(pcoa_plot, legend = "plain") +
  theme(
    axis.line = element_line(linewidth = 1.3, colour = "black"),
    axis.ticks = element_line(linewidth = 1.1, colour = "black"),
    axis.ticks.length = grid::unit(0.22, "cm"),
    axis.title = element_text(size = 34.0, face = "bold", colour = "black"),
    axis.text = element_text(size = 23.0, face = "bold", colour = "black"),
    legend.title = element_text(size = 20.5, face = "bold"),
    legend.text = element_text(size = 18.6),
    legend.key.width = grid::unit(1.0, "cm"),
    legend.key.height = grid::unit(1.1, "cm"),
    legend.spacing.y = grid::unit(0.24, "cm"),
    legend.box = "vertical"
  ) +
  guides(
    fill = guide_legend(
      order = 2,
      override.aes = list(
        shape = 21,
        size = 6.4,
        color = "black",
        stroke = 0.9,
        alpha = 0.95,
        fill = c("Control" = "#B7D9F2", "PS-NP" = "#F2B6A7")
      )
    ),
    shape = guide_legend(
      order = 1,
      override.aes = list(
        fill = "white",
        size = 6.8,
        color = "black",
        stroke = 0.9,
        alpha = 0.95
      )
    )
  )

pcoa_plot
```

# Step 22: Run PERMANOVA Models

```{r}
set.seed(30)
permanova_condition_only <- adonis2(
  bray_dist ~ group_simple,
  data = analysis_metadata,
  permutations = 999
)

set.seed(1)
permanova_batch_only <- adonis2(
  bray_dist ~ accession_number,
  data = analysis_metadata,
  permutations = 999
)

set.seed(3)
permanova_group2_adjusted <- adonis2(
  bray_dist ~ accession_number + group_simple,
  data = analysis_metadata,
  permutations = 999,
  by = "margin"
)

permanova_condition_only
permanova_batch_only
permanova_group2_adjusted
```

# Step 23: Read Genus Relative Abundance Table

```{r}
genus_file <- "/Users/simarjawanda/Library/CloudStorage/OneDrive-UBC/QIIME2 DATA/Group 2/exported-genus-relfreq/genus-relfreq.tsv"

genus_df <- read.delim(genus_file, check.names = FALSE, skip = 1)

dim(genus_df)
head(genus_df[, 1:6])
colnames(genus_df)[1:6]
```

# Step 24: Transpose Genus Table and Merge with Metadata

```{r}
genus_matrix <- genus_df
rownames(genus_matrix) <- genus_matrix[, 1]
genus_matrix <- genus_matrix[, -1]

genus_matrix_t <- as.data.frame(t(genus_matrix))
genus_matrix_t$Run <- rownames(genus_matrix_t)
rownames(genus_matrix_t) <- NULL

genus_merged_df <- analysis_metadata %>%
  inner_join(genus_matrix_t, by = "Run")

dim(genus_merged_df)
```

# Step 25: Run Unadjusted and Stratified Genus Tests

```{r}
genus_abundance_only <- genus_filtered_top50_df %>%
  select(-Run, -group_simple, -psnp_size, -accession_number)

genus_results_unadjusted <- data.frame(
  Genus = character(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

genus_results_stratified <- data.frame(
  Genus = character(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

for (genus_name in colnames(genus_abundance_only)) {
  abundance_values <- as.numeric(genus_abundance_only[[genus_name]])

  test_df_unadjusted <- data.frame(
    group_simple = genus_merged_df$group_simple,
    abundance = abundance_values
  )

  mw <- wilcox.test(abundance ~ group_simple, data = test_df_unadjusted, exact = FALSE)

  genus_results_unadjusted <- rbind(
    genus_results_unadjusted,
    data.frame(
      Genus = genus_name,
      p_value = mw$p.value
    )
  )

  test_df_stratified <- data.frame(
    accession_number = factor(genus_merged_df$accession_number, levels = c("PRJNA944557", "PRJNA1357312")),
    group_simple = factor(genus_merged_df$group_simple, levels = c("control", "psnp")),
    abundance = abundance_values
  )

  stratified_test <- coin::wilcox_test(
    abundance ~ group_simple | accession_number,
    data = test_df_stratified,
    distribution = "asymptotic"
  )

  genus_results_stratified <- rbind(
    genus_results_stratified,
    data.frame(
      Genus = genus_name,
      p_value = as.numeric(coin::pvalue(stratified_test))
    )
  )
}

dim(genus_results_unadjusted)
dim(genus_results_stratified)
head(genus_results_unadjusted)
head(genus_results_stratified)
```

# Step 26: Apply BH/FDR Correction

```{r}
genus_results_unadjusted$q_value <- p.adjust(genus_results_unadjusted$p_value, method = "BH")
genus_results_stratified$q_value <- p.adjust(genus_results_stratified$p_value, method = "BH")

genus_results_unadjusted_raw <- genus_results_unadjusted %>% select(Genus, p_value)
genus_results_stratified_raw <- genus_results_stratified %>% select(Genus, p_value)

genus_results_unadjusted <- genus_results_unadjusted %>%
  arrange(p_value)

genus_results_stratified <- genus_results_stratified %>%
  arrange(p_value)

head(genus_results_unadjusted, 20)
head(genus_results_stratified, 20)

sum(genus_results_unadjusted$q_value < 0.05, na.rm = TRUE)
sum(genus_results_stratified$q_value < 0.05, na.rm = TRUE)
```

# Step 27: Load LEfSe Packages

```{r}
Sys.setenv(
  PATH = paste("/opt/homebrew/bin", Sys.getenv("PATH"), sep = ":"),
  FC = "/opt/homebrew/bin/gfortran",
  F77 = "/opt/homebrew/bin/gfortran",
  F90 = "/opt/homebrew/bin/gfortran"
)

if (!requireNamespace("lefser", quietly = TRUE)) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
  }
  BiocManager::install("lefser", ask = FALSE, update = FALSE)
}

if (!requireNamespace("SummarizedExperiment", quietly = TRUE)) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
  }
  BiocManager::install("SummarizedExperiment", ask = FALSE, update = FALSE)
}

library(lefser)
library(SummarizedExperiment)
library(dplyr)
```

# Step 28: Prepare Genus-Level LEfSe Input

```{r}
lefse_meta <- genus_filtered_top50_df %>%
  select(Run, group_simple, accession_number) %>%
  mutate(
    group_simple = factor(trimws(tolower(group_simple)), levels = c("control", "psnp")),
    accession_number = factor(accession_number, levels = c("PRJNA944557", "PRJNA1357312"))
  ) %>%
  filter(group_simple %in% c("control", "psnp"))

lefse_genus <- genus_filtered_top50_df %>%
  filter(Run %in% lefse_meta$Run) %>%
  mutate(run_order = match(Run, lefse_meta$Run)) %>%
  arrange(run_order) %>%
  select(-run_order, -Run, -group_simple, -psnp_size, -accession_number)

lefse_matrix <- t(as.matrix(lefse_genus))
mode(lefse_matrix) <- "numeric"

lefse_input <- SummarizedExperiment(
  assays = list(relative_abundance = lefse_matrix),
  colData = S4Vectors::DataFrame(lefse_meta)
)

dim(lefse_matrix)
table(lefse_meta$group_simple, lefse_meta$accession_number)
```

# Step 29: Run LEfSe

```{r}
group2_lefse <- lefser(
  lefse_input,
  classCol = "group_simple",
  subclassCol = "accession_number",
  assay = "relative_abundance",
  lda.threshold = 2,
  method = "none",
  trim.names = FALSE
)

group2_lefse <- as.data.frame(group2_lefse)

if (nrow(group2_lefse) > 0) {
  colnames(group2_lefse) <- c("Genus", "LDA_score")
  group2_lefse <- group2_lefse %>%
    arrange(desc(LDA_score))
} else {
  group2_lefse <- data.frame(
    Genus = character(),
    LDA_score = numeric()
  )
}

group2_lefse
```

No significant LEfSe biomarkers were detected for Group 2 under the subclass-aware exploratory setup, so no LDA scores were generated.

# Step 30: Compare LEfSe with the Current Stratified Genus-Level Results

```{r}
group2_lefse_compare <- genus_results_stratified %>%
  mutate(
    Genus = as.character(Genus),
    Found_by_LEfSe = Genus %in% group2_lefse$Genus,
    LEfSe_LDA_score = group2_lefse$LDA_score[match(Genus, group2_lefse$Genus)],
    FDR_significant = q_value < 0.05
  ) %>%
  arrange(p_value)

head(group2_lefse_compare, 20)
sum(group2_lefse_compare$Found_by_LEfSe, na.rm = TRUE)
```

# Step 31: Save LEfSe Outputs

```{r}
lefse_output_dir <- file.path(top50_output_dir, "lefse_group2_outputs")

if (!dir.exists(lefse_output_dir)) {
  dir.create(lefse_output_dir, recursive = TRUE)
}

write.csv(
  group2_lefse,
  file.path(lefse_output_dir, "group2_lefse_results_LDA2.csv"),
  row.names = FALSE
)

write.csv(
  group2_lefse_compare,
  file.path(lefse_output_dir, "group2_lefse_vs_stratified_wilcoxon_results_comparison_LDA2.csv"),
  row.names = FALSE
)

list.files(lefse_output_dir)
```

# Step 32: Recheck Group 2 Top 50 Genus Test Tables Before Saving

```{r}
expected_top50_genus_n <- ncol(genus_filtered_top50_df) - 4

if (nrow(genus_results_unadjusted) != expected_top50_genus_n ||
    nrow(genus_results_stratified) != expected_top50_genus_n) {

  genus_abundance_only <- genus_filtered_top50_df %>%
    select(-Run, -group_simple, -psnp_size, -accession_number)

  genus_results_unadjusted <- data.frame(
    Genus = character(),
    p_value = numeric(),
    stringsAsFactors = FALSE
  )

  genus_results_stratified <- data.frame(
    Genus = character(),
    p_value = numeric(),
    stringsAsFactors = FALSE
  )

  for (genus_name in colnames(genus_abundance_only)) {
    abundance_values <- as.numeric(genus_abundance_only[[genus_name]])

    test_df_unadjusted <- data.frame(
      group_simple = genus_filtered_top50_df$group_simple,
      abundance = abundance_values
    )

    mw <- wilcox.test(abundance ~ group_simple, data = test_df_unadjusted, exact = FALSE)

    genus_results_unadjusted <- rbind(
      genus_results_unadjusted,
      data.frame(
        Genus = genus_name,
        p_value = mw$p.value
      )
    )

    test_df_stratified <- data.frame(
      accession_number = factor(genus_filtered_top50_df$accession_number, levels = c("PRJNA944557", "PRJNA1357312")),
      group_simple = factor(genus_filtered_top50_df$group_simple, levels = c("control", "psnp")),
      abundance = abundance_values
    )

    stratified_test <- coin::wilcox_test(
      abundance ~ group_simple | accession_number,
      data = test_df_stratified,
      distribution = "asymptotic"
    )

    genus_results_stratified <- rbind(
      genus_results_stratified,
      data.frame(
        Genus = genus_name,
        p_value = as.numeric(coin::pvalue(stratified_test))
      )
    )
  }

  genus_results_unadjusted$q_value <- p.adjust(genus_results_unadjusted$p_value, method = "BH")
  genus_results_stratified$q_value <- p.adjust(genus_results_stratified$p_value, method = "BH")

  genus_results_unadjusted <- genus_results_unadjusted %>%
    arrange(p_value)

  genus_results_stratified <- genus_results_stratified %>%
    arrange(p_value)

  genus_results_unadjusted_raw <- genus_results_unadjusted %>% select(Genus, p_value)
  genus_results_stratified_raw <- genus_results_stratified %>% select(Genus, p_value)
}

dim(genus_results_unadjusted)
dim(genus_results_stratified)
```

# Step 33: Build Taxa Plotting Data

```{r}
genus_plot_df <- genus_merged_df %>%
  select(-psnp_size) %>%
  pivot_longer(
    cols = -c(Run, group_simple, accession_number),
    names_to = "Taxonomy",
    values_to = "RelativeAbundance"
  ) %>%
  mutate(
    RelativeAbundance = as.numeric(RelativeAbundance),
    accession_number = factor(accession_number, levels = c("PRJNA944557", "PRJNA1357312")),
    Condition = ifelse(group_simple == "psnp", "PS-NP", "Control"),
    Condition = factor(Condition, levels = c("Control", "PS-NP")),
    Genus = clean_taxon_label_composition(Taxonomy)
  )

head(genus_plot_df)
```

# Step 34: Make Top 10 Taxa Barplot

```{r top10_taxa_plot_group2, fig.width=11.5, fig.height=8.8}
top10_genera <- genus_plot_df %>%
  group_by(Genus) %>%
  summarise(TotalAbundance = sum(RelativeAbundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(TotalAbundance)) %>%
  dplyr::slice(1:10) %>%
  pull(Genus)

genus_plot_top10 <- genus_plot_df %>%
  filter(Genus %in% top10_genera) %>%
  group_by(Condition, Genus) %>%
  summarise(
    MeanRelativeAbundance = mean(RelativeAbundance, na.rm = TRUE),
    .groups = "drop"
  )

genus_order <- genus_plot_top10 %>%
  group_by(Genus) %>%
  summarise(TotalAbundance = sum(MeanRelativeAbundance), .groups = "drop") %>%
  arrange(desc(TotalAbundance)) %>%
  pull(Genus)

stack_order <- genus_order
legend_order <- rev(stack_order)

genus_plot_top10$Genus <- factor(genus_plot_top10$Genus, levels = stack_order)

genus_palette <- build_shared_genus_palette_composition(stack_order)

top10_taxa_plot <- ggplot(genus_plot_top10, aes(x = Condition, y = MeanRelativeAbundance, fill = Genus)) +
  geom_col(
    position = position_fill(reverse = TRUE),
    width = 0.55,
    color = "white",
    linewidth = 0.08
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = genus_palette,
    breaks = legend_order
  ) +
  labs(
    title = wrap_figure_title("Ten Most Abundant Genera by Mean Relative Abundance in Control and PS-NP Samples for Group 2"),
    x = NULL,
    y = "Mean Relative Abundance (%)",
    fill = "Genus"
  ) +
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(size = 17, hjust = 0.5, face = "plain", lineheight = 1.05, margin = margin(b = 10)),
    axis.title.y = element_text(size = 15),
    axis.text.x = element_text(size = 14, color = "black"),
    axis.text.y = element_text(size = 13, color = "black"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 9.5, face = "italic"),
    legend.position = "right",
    legend.key.height = unit(0.34, "cm"),
    legend.spacing.y = unit(0.08, "cm")
  )

top10_taxa_plot <- finalize_thesis_plot(top10_taxa_plot, legend = "italic")

top10_taxa_plot
```

# Step 35: Make All-Genera Supplementary Taxa Barplot

```{r all_genera_plot_group2, fig.width=12, fig.height=8}
clean_taxon_label <- function(x) clean_taxon_label_composition(x)

genus_plot_all <- genus_plot_df %>%
  mutate(
    Genus = clean_taxon_label(Taxonomy)
  ) %>%
  group_by(Condition, Genus) %>%
  summarise(
    MeanRelativeAbundance = mean(RelativeAbundance, na.rm = TRUE),
    .groups = "drop"
  )

genus_order_all <- genus_plot_all %>%
  group_by(Genus) %>%
  summarise(TotalAbundance = sum(MeanRelativeAbundance), .groups = "drop") %>%
  arrange(desc(TotalAbundance)) %>%
  pull(Genus)

stack_order_all <- genus_order_all
legend_order_all <- rev(stack_order_all)

genus_plot_all$Genus <- factor(genus_plot_all$Genus, levels = stack_order_all)

composition_palette <- build_shared_genus_palette_composition(stack_order_all)

all_genera_plot <- ggplot(genus_plot_all, aes(x = Condition, y = MeanRelativeAbundance, fill = Genus)) +
  geom_col(
    position = position_fill(reverse = TRUE),
    width = 0.55,
    color = "white",
    linewidth = 0.05
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = composition_palette,
    breaks = legend_order_all
  ) +
  labs(
    title = wrap_figure_title("Mean Relative Abundance of All Detected Genera in Control and PS-NP Samples for Group 2"),
    x = NULL,
    y = "Mean Relative Abundance (%)",
    fill = "Genus"
  ) +
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(size = 17, hjust = 0.5, face = "plain", lineheight = 1.05, margin = margin(b = 10)),
    axis.title.y = element_text(size = 15),
    axis.text.x = element_text(size = 14, color = "black"),
    axis.text.y = element_text(size = 13, color = "black"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 8.5, face = "italic"),
    legend.position = "right",
    legend.key.height = unit(0.24, "cm"),
    legend.spacing.y = unit(0.05, "cm")
  ) +
  guides(fill = guide_legend(ncol = 2))

all_genera_plot <- finalize_thesis_plot(all_genera_plot, legend = "italic") +
  theme(
    legend.title = element_text(size = 13.5, face = "bold"),
    legend.text = element_text(size = 9.5, face = "italic"),
    legend.key.height = unit(0.20, "cm"),
    legend.key.width = unit(0.24, "cm"),
    legend.spacing.y = unit(0.02, "cm"),
    legend.box.spacing = unit(0.08, "cm")
  ) +
  guides(fill = guide_legend(ncol = 2, byrow = TRUE))

all_genera_plot
```

# Step 36: Make Per-Sample Taxa Barplot

```{r per_sample_taxa_plot_group2, fig.width=15.5, fig.height=9.2}
sample_order_df <- genus_plot_df %>%
  distinct(accession_number, Condition, Run) %>%
  mutate(
    accession_number = factor(accession_number, levels = c("PRJNA944557", "PRJNA1357312")),
    Condition = factor(Condition, levels = c("Control", "PS-NP"))
  ) %>%
  arrange(accession_number, Condition, Run)

genus_plot_sample <- genus_plot_df %>%
  filter(Genus %in% top10_genera) %>%
  group_by(Run, accession_number, Condition, Genus) %>%
  summarise(
    RelativeAbundance = sum(RelativeAbundance, na.rm = TRUE),
    .groups = "drop"
  )

genus_plot_sample$Run <- factor(genus_plot_sample$Run, levels = sample_order_df$Run)
genus_plot_sample$Genus <- factor(genus_plot_sample$Genus, levels = stack_order)

per_sample_taxa_plot <- ggplot(genus_plot_sample, aes(x = Run, y = RelativeAbundance, fill = Genus)) +
  geom_col(
    position = position_fill(reverse = TRUE),
    width = 0.9,
    color = "white",
    linewidth = 0.05
  ) +
  ggh4x::facet_nested(
    ~ accession_number + Condition,
    scales = "free_x",
    space = "free_x"
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = genus_palette,
    breaks = legend_order
  ) +
  labs(
    title = wrap_figure_title("Relative Abundance of the Ten Most Abundant Genera Across Individual Samples in Group 2, Grouped by Study and Condition"),
    x = NULL,
    y = "Relative Abundance (%)",
    fill = "Genus"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5, face = "plain", lineheight = 1.05, margin = margin(b = 10)),
    axis.title = element_text(size = 15),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8, color = "black"),
    axis.text.y = element_text(size = 12, color = "black"),
    strip.background = element_blank(),
    strip.text = element_text(size = 12, face = "plain"),
    ggh4x.facet.nestline = element_blank(),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 9, face = "italic"),
    legend.position = "right",
    legend.key.height = unit(0.28, "cm"),
    legend.spacing.y = unit(0.06, "cm")
  )

per_sample_taxa_plot <- finalize_thesis_plot(per_sample_taxa_plot, legend = "italic") +
  theme(
    axis.title.y = element_text(size = 17, face = "bold", colour = "black"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 9, face = "bold", colour = "black"),
    axis.text.y = element_text(size = 13.5, face = "bold", colour = "black"),
    strip.background = element_blank(),
    strip.text = element_text(size = 13.5, face = "bold", colour = "black"),
    ggh4x.facet.nestline = element_blank()
  )

per_sample_taxa_plot
```

# Step 37: Run Per-Study Exploratory Checks for Group 2 Later

```{r eval=FALSE}
study_ids <- c("PRJNA944557", "PRJNA1357312")

sanity_alpha_results <- data.frame(
  accession_number = character(),
  metric = character(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

sanity_permanova_results <- data.frame(
  accession_number = character(),
  Df = numeric(),
  SumOfSqs = numeric(),
  R2 = numeric(),
  F = numeric(),
  Pr = numeric(),
  stringsAsFactors = FALSE
)

sanity_genus_results <- data.frame(
  accession_number = character(),
  Genus = character(),
  p_value = numeric(),
  q_value = numeric(),
  stringsAsFactors = FALSE
)

for (study_id in study_ids) {
  study_meta <- analysis_metadata %>%
    filter(accession_number == study_id)

  study_alpha <- alpha_meta_df %>%
    filter(accession_number == study_id)

  mw_shannon <- wilcox.test(Shannon ~ group_simple, data = study_alpha, exact = FALSE)
  mw_simpson <- wilcox.test(Simpson ~ group_simple, data = study_alpha, exact = FALSE)

  sanity_alpha_results <- rbind(
    sanity_alpha_results,
    data.frame(accession_number = study_id, metric = "Shannon", p_value = mw_shannon$p.value),
    data.frame(accession_number = study_id, metric = "Simpson", p_value = mw_simpson$p.value)
  )

  study_norm_abundance <- norm_abundance_df[rownames(norm_abundance_df) %in% study_meta$Run, , drop = FALSE]
  study_norm_abundance <- study_norm_abundance[study_meta$Run, , drop = FALSE]

  study_bray <- vegdist(study_norm_abundance, method = "bray")
  if (study_id == "PRJNA1357312") {
    set.seed(7)
  } else if (study_id == "PRJNA944557") {
    set.seed(12)
  }
  study_permanova <- adonis2(study_bray ~ group_simple, data = study_meta, permutations = 999)

  sanity_permanova_results <- rbind(
    sanity_permanova_results,
    data.frame(
      accession_number = study_id,
      Df = study_permanova$Df[1],
      SumOfSqs = study_permanova$SumOfSqs[1],
      R2 = study_permanova$R2[1],
      F = study_permanova$F[1],
      Pr = study_permanova$`Pr(>F)`[1]
    )
  )

  study_genus <- genus_merged_df %>%
    filter(accession_number == study_id)

  study_genus_abundance <- study_genus %>%
    select(-Run, -group_simple, -psnp_size, -accession_number)

  study_genus_res <- data.frame(
    accession_number = character(),
    Genus = character(),
    p_value = numeric(),
    stringsAsFactors = FALSE
  )

  for (genus_name in colnames(study_genus_abundance)) {
    test_df <- data.frame(
      group_simple = study_genus$group_simple,
      abundance = as.numeric(study_genus_abundance[[genus_name]])
    )

    mw <- wilcox.test(abundance ~ group_simple, data = test_df, exact = FALSE)

    study_genus_res <- rbind(
      study_genus_res,
      data.frame(
        accession_number = study_id,
        Genus = genus_name,
        p_value = mw$p.value
      )
    )
  }

  study_genus_res$q_value <- p.adjust(study_genus_res$p_value, method = "BH")
  study_genus_res <- study_genus_res %>% arrange(p_value)

  sanity_genus_results <- rbind(sanity_genus_results, study_genus_res)
}

sanity_alpha_results
sanity_permanova_results
head(sanity_genus_results, 20)
```

# Step 38: Save All Outputs

```{r}
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

group2_saved_genus_unadjusted <- data.frame(
  Genus = character(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

group2_saved_genus_stratified <- data.frame(
  Genus = character(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

group2_saved_filtered_top50_df <- read.csv(
  file.path(top50_output_dir, "group2_genus_filtered_top50_for_downstream.csv"),
  check.names = FALSE
)

group2_saved_genus_abundance <- group2_saved_filtered_top50_df %>%
  select(-Run, -group_simple, -psnp_size, -accession_number)

for (genus_name in colnames(group2_saved_genus_abundance)) {
  abundance_values <- as.numeric(group2_saved_genus_abundance[[genus_name]])

  test_df_unadjusted <- data.frame(
    group_simple = group2_saved_filtered_top50_df$group_simple,
    abundance = abundance_values
  )

  mw <- wilcox.test(abundance ~ group_simple, data = test_df_unadjusted, exact = FALSE)

  group2_saved_genus_unadjusted <- rbind(
    group2_saved_genus_unadjusted,
    data.frame(
      Genus = genus_name,
      p_value = mw$p.value
    )
  )

  test_df_stratified <- data.frame(
    accession_number = factor(group2_saved_filtered_top50_df$accession_number, levels = c("PRJNA944557", "PRJNA1357312")),
    group_simple = factor(group2_saved_filtered_top50_df$group_simple, levels = c("control", "psnp")),
    abundance = abundance_values
  )

  stratified_test <- coin::wilcox_test(
    abundance ~ group_simple | accession_number,
    data = test_df_stratified,
    distribution = "asymptotic"
  )

  group2_saved_genus_stratified <- rbind(
    group2_saved_genus_stratified,
    data.frame(
      Genus = genus_name,
      p_value = as.numeric(coin::pvalue(stratified_test))
    )
  )
}

group2_saved_genus_unadjusted$q_value <- p.adjust(group2_saved_genus_unadjusted$p_value, method = "BH")
group2_saved_genus_stratified$q_value <- p.adjust(group2_saved_genus_stratified$p_value, method = "BH")

group2_saved_genus_unadjusted <- group2_saved_genus_unadjusted %>%
  arrange(p_value)

group2_saved_genus_stratified <- group2_saved_genus_stratified %>%
  arrange(p_value)

group2_saved_genus_unadjusted_raw <- group2_saved_genus_unadjusted %>% select(Genus, p_value)
group2_saved_genus_stratified_raw <- group2_saved_genus_stratified %>% select(Genus, p_value)

write.csv(analysis_metadata, file.path(output_dir, "group2_analysis_metadata.csv"), row.names = FALSE)
write.csv(group_summary_df, file.path(output_dir, "group2_group_summary.csv"), row.names = FALSE)

write.csv(abundance_df, file.path(output_dir, "group2_abundance_cleaned.csv"), row.names = FALSE)
write.csv(as.data.frame(norm_abundance_df), file.path(output_dir, "group2_abundance_normalized.csv"), row.names = TRUE)

write.csv(alpha_df, file.path(output_dir, "group2_alpha_diversity.csv"), row.names = FALSE)
write.csv(alpha_meta_df, file.path(output_dir, "group2_alpha_diversity_with_metadata.csv"), row.names = FALSE)
write.csv(shapiro_shannon, file.path(output_dir, "group2_shapiro_shannon_by_condition.csv"), row.names = FALSE)
write.csv(shapiro_simpson, file.path(output_dir, "group2_shapiro_simpson_by_condition.csv"), row.names = FALSE)
write.csv(broom::tidy(mann_whitney_shannon), file.path(output_dir, "group2_mann_whitney_shannon.csv"), row.names = FALSE)
write.csv(broom::tidy(mann_whitney_simpson), file.path(output_dir, "group2_mann_whitney_simpson.csv"), row.names = FALSE)
write.csv(stratified_wilcoxon_shannon, file.path(output_dir, "group2_shannon_stratified_wilcoxon.csv"), row.names = FALSE)
write.csv(stratified_wilcoxon_simpson, file.path(output_dir, "group2_simpson_stratified_wilcoxon.csv"), row.names = FALSE)

write.csv(as.data.frame(as.matrix(bray_dist)), file.path(output_dir, "group2_bray_curtis_distance_matrix.csv"), row.names = TRUE)
write.csv(pcoa_plot_df, file.path(output_dir, "group2_pcoa_coordinates.csv"), row.names = FALSE)
write.csv(as.data.frame(permanova_condition_only), file.path(output_dir, "group2_permanova_condition_only.csv"), row.names = TRUE)
write.csv(as.data.frame(permanova_batch_only), file.path(output_dir, "group2_permanova_batch_only.csv"), row.names = TRUE)
write.csv(as.data.frame(permanova_group2_adjusted), file.path(output_dir, "group2_permanova_batch_adjusted.csv"), row.names = TRUE)

write.csv(
  group2_saved_genus_unadjusted_raw,
  file.path(top50_output_dir, "group2_genus_level_mann_whitney_raw_p_values_top50.csv"),
  row.names = FALSE
)

write.csv(
  group2_saved_genus_unadjusted,
  file.path(top50_output_dir, "group2_genus_level_mann_whitney_fdr_results_top50.csv"),
  row.names = FALSE
)

write.csv(
  group2_saved_genus_stratified_raw,
  file.path(top50_output_dir, "group2_genus_level_stratified_wilcoxon_raw_p_values_top50.csv"),
  row.names = FALSE
)

write.csv(
  group2_saved_genus_stratified,
  file.path(top50_output_dir, "group2_genus_level_stratified_wilcoxon_fdr_results_top50.csv"),
  row.names = FALSE
)

write.csv(genus_plot_top10, file.path(output_dir, "group2_top10_taxa_barplot_data.csv"), row.names = FALSE)
write.csv(genus_plot_all, file.path(output_dir, "group2_all_genera_barplot_data.csv"), row.names = FALSE)
write.csv(genus_plot_sample, file.path(output_dir, "group2_per_sample_top10_taxa_barplot_data.csv"), row.names = FALSE)

ggsave(file.path(output_dir, "group2_shannon_boxplot.png"), plot = shannon_plot, width = 8.4, height = 6.8, dpi = 600)
ggsave(file.path(output_dir, "group2_simpson_boxplot.png"), plot = simpson_plot, width = 8.4, height = 6.8, dpi = 600)
ggsave(file.path(output_dir, "group2_pcoa_plot.png"), plot = pcoa_plot, width = 8.4, height = 6.7, dpi = 600)
ggsave(file.path(output_dir, "group2_top10_taxa_barplot.png"), plot = top10_taxa_plot, width = 11.5, height = 8.8, dpi = 600)
ggsave(file.path(output_dir, "group2_all_genera_supplementary_barplot.png"), plot = all_genera_plot, width = 16.2, height = 9.8, dpi = 600)
ggsave(file.path(output_dir, "group2_per_sample_top10_taxa_barplot.png"), plot = per_sample_taxa_plot, width = 15.5, height = 9.2, dpi = 600)

cat("All Group 2 outputs saved to:\n", output_dir, "\n")
print(list.files(output_dir))
```

# Step 39: Select Candidate Genera for Follow-Up Plots

```{r}
clean_taxon_label_followup <- function(x) {
  genus <- extract_taxonomy_rank(x, "g")
  family <- extract_taxonomy_rank(x, "f")
  order <- extract_taxonomy_rank(x, "o")
  class_name <- extract_taxonomy_rank(x, "c")
  phylum <- extract_taxonomy_rank(x, "p")

  genus_invalid <- genus == "" | is.na(genus) | grepl("^[a-z]__$", genus)
  parent_label <- build_parent_rank_label(family, order, class_name, phylum)

  label <- ifelse(
    tolower(genus) == "uncultured" & !is.na(parent_label),
    paste0("Uncultured (", parent_label, ")"),
    ifelse(
      tolower(genus) %in% c("unassigned", "unclassified", "unknown", "incertae sedis") & !is.na(parent_label),
      paste0(tools::toTitleCase(genus), " (", parent_label, ")"),
      ifelse(
        !genus_invalid,
        genus,
        ifelse(!is.na(parent_label), paste0("Unassigned (", parent_label, ")"), "Unassigned (taxonomy unavailable)")
      )
    )
  )

  label
}

build_shared_genus_palette_followup <- function(genus_names) {
  genus_palette <- c(
    "Lactobacillus" = "#8ECFC9",
    "Muribaculaceae" = "#F3C78D",
    "Streptococcus" = "#A8B6D9",
    "Turicibacter" = "#E8A0A6",
    "Faecalibaculum" = "#A9CFA5",
    "Odoribacter" = "#DCC7A1",
    "Bifidobacterium" = "#C8B6D8",
    "Rikenellaceae RC9 gut group" = "#E7B8D6",
    "Alistipes" = "#90B4C6",
    "Helicobacter" = "#D8A48F",
    "Akkermansia" = "#B7A0D8",
    "Bacteroides" = "#8FBE6D",
    "Clostridia UCG-014" = "#B48A8A",
    "Clostridium sensu stricto 1" = "#8D6CCF",
    "Lachnospiraceae NK4A136 group" = "#5B84B1",
    "Muribaculum" = "#9E9D68",
    "Candidatus Saccharimonas" = "#6F8FB4",
    "Desulfovibrio" = "#A77C95",
    "Dubosiella" = "#8AA24A",
    "Enterobacter" = "#E76F51",
    "Blautia" = "#1F78B4",
    "Klebsiella" = "#F39C12",
    "Parasutterella" = "#2F9C95",
    "Candidatus Arthromitus" = "#8E44AD",
    "Incertae Sedis" = "#5FAE8D",
    "Incertae Sedis (Ruminococcaceae)" = "#5FAE8D",
    "RF39" = "#4A90D9",
    "Rikenella" = "#A26CC5",
    "Prevotellaceae UCG-001" = "#2E8B57",
    "[Clostridium] innocuum group" = "#CC8456",
    "Lachnospiraceae UCG-001" = "#5B89C9",
    "[Eubacterium] ventriosum group" = "#C46A93",
    "Uncultured" = "#7BAFD4",
    "Uncultured (Lachnospiraceae)" = "#9FB6C8",
    "Uncultured (Ruminococcaceae)" = "#AFC1D8",
    "Uncultured (Oscillospiraceae)" = "#8AAFD0",
    "Unassigned" = "#C892A6",
    "Unassigned (taxonomy unavailable)" = "#C892A6",
    "Unassigned (Lachnospiraceae)" = "#E0B180",
    "Unclassified" = "#C88EBE",
    "Unclassified (Lachnospiraceae)" = "#D5B3C6",
    "Unclassified (Clostridia)" = "#CFA4BF",
    "Unclassified (Erysipelatoclostridiaceae)" = "#DAB7C5"
  )

  missing_genera <- setdiff(genus_names, names(genus_palette))

  if (length(missing_genera) > 0) {
    extra_cols <- deterministic_extra_colors(missing_genera)
    genus_palette <- c(genus_palette, extra_cols)
  }

  genus_palette
}

select_candidate_genera_followup <- function(results_df, max_raw = 8) {
  results_df <- results_df %>%
    mutate(
      DisplayGenus = clean_taxon_label_followup(Genus)
    ) %>%
    arrange(p_value)

  fdr_hits <- results_df %>%
    filter(q_value < 0.05)

  raw_hits <- results_df %>%
    filter(p_value < 0.05)

  if (nrow(fdr_hits) > 0) {
    selected_df <- fdr_hits %>%
      mutate(selection_reason = "FDR_significant")
  } else if (nrow(raw_hits) > 0) {
    selected_df <- raw_hits %>%
      slice_head(n = max_raw) %>%
      mutate(selection_reason = "Raw_p_below_0.05")
  } else {
    selected_df <- results_df %>%
      slice_head(n = 3) %>%
      mutate(selection_reason = "Top_raw_p_values")
  }

  selected_df
}

group2_followup_filtered_top50_df <- read.csv(
  file.path(top50_output_dir, "group2_genus_filtered_top50_for_downstream.csv"),
  check.names = FALSE
)

group2_followup_genus_abundance <- group2_followup_filtered_top50_df %>%
  select(-Run, -group_simple, -psnp_size, -accession_number)

group2_followup_results_stratified <- data.frame(
  Genus = character(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

for (genus_name in colnames(group2_followup_genus_abundance)) {
  abundance_values <- as.numeric(group2_followup_genus_abundance[[genus_name]])

  test_df_stratified <- data.frame(
    accession_number = factor(group2_followup_filtered_top50_df$accession_number, levels = c("PRJNA944557", "PRJNA1357312")),
    group_simple = factor(group2_followup_filtered_top50_df$group_simple, levels = c("control", "psnp")),
    abundance = abundance_values
  )

  stratified_test <- coin::wilcox_test(
    abundance ~ group_simple | accession_number,
    data = test_df_stratified,
    distribution = "asymptotic"
  )

  group2_followup_results_stratified <- rbind(
    group2_followup_results_stratified,
    data.frame(
      Genus = genus_name,
      p_value = as.numeric(coin::pvalue(stratified_test))
    )
  )
}

group2_followup_results_stratified$q_value <- p.adjust(group2_followup_results_stratified$p_value, method = "BH")
group2_followup_results_stratified <- group2_followup_results_stratified %>%
  arrange(p_value)

selected_genus_candidates <- select_candidate_genera_followup(group2_followup_results_stratified, max_raw = 8) %>%
  mutate(test_type = "stratified_wilcoxon")

selected_genus_boxplot_data <- genus_merged_df %>%
  select(Run, group_simple, accession_number, all_of(selected_genus_candidates$Genus)) %>%
  pivot_longer(
    cols = -c(Run, group_simple, accession_number),
    names_to = "Genus",
    values_to = "RelativeAbundance"
  ) %>%
  mutate(
    RelativeAbundance = as.numeric(RelativeAbundance),
    Condition = ifelse(group_simple == "psnp", "PS-NP", "Control"),
    Condition = factor(Condition, levels = c("Control", "PS-NP")),
    accession_number = factor(accession_number, levels = c("PRJNA944557", "PRJNA1357312")),
    DisplayGenus = clean_taxon_label_followup(Genus)
  ) %>%
  left_join(
    selected_genus_candidates %>%
      select(Genus, DisplayGenus, p_value, q_value, selection_reason, test_type),
    by = c("Genus", "DisplayGenus")
  )

selected_genus_order <- selected_genus_boxplot_data %>%
  group_by(DisplayGenus) %>%
  summarise(TotalAbundance = mean(RelativeAbundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(TotalAbundance)) %>%
  pull(DisplayGenus)

selected_genus_boxplot_data$DisplayGenus <- factor(
  selected_genus_boxplot_data$DisplayGenus,
  levels = selected_genus_order
)

selected_genus_barplot_data <- selected_genus_boxplot_data %>%
  group_by(Condition, DisplayGenus) %>%
  summarise(
    MeanRelativeAbundance = mean(RelativeAbundance, na.rm = TRUE),
    p_value = dplyr::first(p_value),
    q_value = dplyr::first(q_value),
    selection_reason = dplyr::first(selection_reason),
    test_type = dplyr::first(test_type),
    .groups = "drop"
  )

selected_genus_stack_order <- selected_genus_barplot_data %>%
  group_by(DisplayGenus) %>%
  summarise(TotalAbundance = sum(MeanRelativeAbundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(TotalAbundance)) %>%
  pull(DisplayGenus)

selected_genus_barplot_data$DisplayGenus <- factor(
  selected_genus_barplot_data$DisplayGenus,
  levels = selected_genus_stack_order
)

selected_genus_palette <- build_shared_genus_palette_followup(levels(selected_genus_barplot_data$DisplayGenus))

selected_genus_annotation <- selected_genus_boxplot_data %>%
  group_by(DisplayGenus) %>%
  summarise(
    p_value = dplyr::first(p_value),
    q_value = dplyr::first(q_value),
    label = dplyr::case_when(
      q_value < 0.05 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ ""
    ),
    .groups = "drop"
  )

selected_genus_panel_map <- make_panel_label_map(levels(selected_genus_boxplot_data$DisplayGenus))

selected_genus_boxplot_data$DisplayGenusPanel <- factor(
  selected_genus_panel_map[as.character(selected_genus_boxplot_data$DisplayGenus)],
  levels = unname(selected_genus_panel_map)
)

selected_genus_annotation$DisplayGenusPanel <- factor(
  selected_genus_panel_map[as.character(selected_genus_annotation$DisplayGenus)],
  levels = unname(selected_genus_panel_map)
)

selected_genus_brackets <- build_two_group_brackets(
  selected_genus_boxplot_data %>%
    left_join(selected_genus_annotation %>% select(DisplayGenus, label), by = "DisplayGenus"),
  facet_col = "DisplayGenusPanel",
  value_col = "RelativeAbundance",
  label_col = "label"
)

selected_genus_panel_groups <- split(
  levels(selected_genus_boxplot_data$DisplayGenusPanel),
  ceiling(seq_along(levels(selected_genus_boxplot_data$DisplayGenusPanel)) / 4)
)

selected_genus_candidates
```

# Step 40: Make Candidate Genus Abundance Boxplots

```{r group2_selected_genus_boxplots, fig.width=18.4, fig.height=13.8}
selected_genus_boxplot <- make_selected_genus_panel_grid(
  selected_genus_boxplot_data,
  selected_genus_brackets,
  selected_genus_palette,
  levels(selected_genus_boxplot_data$DisplayGenusPanel),
  ncol = 2
)
selected_genus_boxplot
```

# Step 41: Make Candidate Genus Mean-Abundance Barplot

```{r group2_selected_genus_barplot, fig.width=14.0, fig.height=9.6}
selected_genus_barplot <- ggplot(
  selected_genus_barplot_data,
  aes(x = Condition, y = MeanRelativeAbundance, fill = DisplayGenus)
) +
  geom_col(
    position = position_stack(reverse = TRUE),
    width = 0.68,
    color = "white",
    linewidth = 0.08
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 0.1),
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = selected_genus_palette,
    breaks = rev(levels(selected_genus_barplot_data$DisplayGenus))
  ) +
  labs(
    title = wrap_figure_title("Mean Relative Abundance of Candidate Genera Differing Between Control and PS-NP Samples in Group 2", width = 56),
    x = "Condition",
    y = "Mean Relative Abundance (%)",
    fill = "Genus"
  ) +
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(size = 18, hjust = 0.5, face = "plain", lineheight = 1.05, margin = margin(b = 10)),
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 12, color = "black"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 9.5, face = "italic"),
    legend.position = "right",
    plot.title.position = "plot",
    plot.margin = margin(t = 10, r = 16, b = 12, l = 16),
    legend.key.height = unit(0.3, "cm"),
    legend.spacing.y = unit(0.06, "cm")
  )

selected_genus_barplot <- finalize_thesis_plot(selected_genus_barplot, legend = "italic")

selected_genus_barplot
```

# Step 42: Save Candidate Genus Follow-Up Outputs

```{r}
write.csv(
  selected_genus_candidates,
  file.path(output_dir, "group2_selected_genus_candidates_top50.csv"),
  row.names = FALSE
)

write.csv(
  selected_genus_boxplot_data,
  file.path(output_dir, "group2_selected_genus_boxplot_data_top50.csv"),
  row.names = FALSE
)

write.csv(
  selected_genus_barplot_data,
  file.path(output_dir, "group2_selected_genus_barplot_data_top50.csv"),
  row.names = FALSE
)

old_group2_part_files <- list.files(
  output_dir,
  pattern = "^group2_selected_genus_boxplots_top50_part[0-9]+\\.png$",
  full.names = TRUE
)

if (length(old_group2_part_files) > 0) {
  file.remove(old_group2_part_files)
}

for (idx in seq_along(selected_genus_panel_groups)) {
  chunk_plot <- make_selected_genus_panel_grid(
    selected_genus_boxplot_data,
    selected_genus_brackets,
    selected_genus_palette,
    selected_genus_panel_groups[[idx]],
    ncol = 2
  )
  chunk_rows <- ceiling(length(selected_genus_panel_groups[[idx]]) / 2)
  chunk_height <- ifelse(chunk_rows == 1, 8.4, 14.0)

  ggsave(
    file.path(output_dir, sprintf("group2_selected_genus_boxplots_top50_part%d.png", idx)),
    plot = chunk_plot,
    width = 18.4,
    height = chunk_height,
    dpi = 600
  )
}

ggsave(
  file.path(output_dir, "group2_selected_genus_boxplots_top50.png"),
  plot = selected_genus_boxplot,
  width = 18.4,
  height = 25.5,
  dpi = 600
)

ggsave(
  file.path(output_dir, "group2_selected_genus_barplot_top50.png"),
  plot = selected_genus_barplot,
  width = 14.0,
  height = 9.6,
  dpi = 600
)
```
