---
title: "Group 2 Exploratory Analysis Differentiated by BioProject Accession"
author: "Simar Jawanda"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  dev = "png",
  fig.ext = "png",
  dpi = 300
)
```

# Step 1: Load Packages

```{r}
library(broom)
library(dplyr)
library(ggh4x)
library(ggplot2)
library(tidyr)
library(tibble)
library(vegan)
library(scales)

wrap_figure_title <- function(title_text, width = 62) {
  paste(strwrap(title_text, width = width), collapse = "\n")
}

finalize_thesis_plot <- function(plot, legend = c("none", "plain", "italic")) {
  legend <- match.arg(legend)

  if (!is.null(plot$labels$x) && identical(plot$labels$x, "Condition")) {
    plot <- plot + labs(x = NULL)
  }

  plot <- plot +
    labs(title = NULL) +
    theme(
      plot.title = element_blank(),
      plot.title.position = "panel",
      axis.line = element_line(linewidth = 1.1, colour = "black"),
      axis.ticks = element_line(linewidth = 0.95, colour = "black"),
      axis.ticks.length = grid::unit(0.18, "cm"),
      axis.title.x = element_text(size = 15, face = "bold", colour = "black"),
      axis.title.y = element_text(size = 15, face = "bold", colour = "black"),
      axis.text.x = element_text(size = 12.5, face = "bold", colour = "black"),
      axis.text.y = element_text(size = 12.5, face = "bold", colour = "black"),
      strip.background = element_rect(fill = "white", colour = "black", linewidth = 1.1),
      strip.text = element_text(size = 13.5, face = "bold", colour = "black")
    )

  if (legend != "none") {
    plot <- plot +
      guides(
        fill = guide_legend(ncol = 1, byrow = TRUE),
        colour = guide_legend(ncol = 1, byrow = TRUE),
        shape = guide_legend(ncol = 1, byrow = TRUE)
      ) +
      theme(
        legend.title = element_text(size = 17, face = "bold"),
        legend.text = element_text(size = 14.5, face = ifelse(legend == "italic", "italic", "plain")),
        legend.key.width = grid::unit(0.55, "cm"),
        legend.key.height = grid::unit(0.62, "cm"),
        legend.spacing.y = grid::unit(0.14, "cm"),
        legend.box = "vertical"
      )
  }

  plot
}

make_panel_label_map <- function(labels) {
  panel_labels <- paste0(LETTERS[seq_along(labels)], ". ", labels)
  stats::setNames(panel_labels, labels)
}

build_two_group_brackets <- function(data, facet_col, value_col, label_col) {
  data %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(facet_col))) %>%
    dplyr::summarise(
      bracket_y = max(.data[[value_col]], na.rm = TRUE) * 1.20 + 1e-06,
      label = dplyr::first(.data[[label_col]]),
      .groups = "drop"
    ) %>%
    dplyr::filter(!is.na(label), label != "") %>%
    dplyr::mutate(
      x_start = 1,
      x_end = 2,
      x_mid = 1.5,
      tick_drop = pmax(bracket_y * 0.04, 1e-06),
      label_y = bracket_y + pmax(bracket_y * 0.055, 1e-06)
    )
}

add_two_group_brackets <- function(plot, bracket_df, text_size = 6) {
  if (nrow(bracket_df) == 0) {
    return(plot)
  }

  plot +
    geom_segment(
      data = bracket_df,
      aes(x = x_start, xend = x_end, y = bracket_y, yend = bracket_y),
      inherit.aes = FALSE,
      linewidth = 0.9
    ) +
    geom_segment(
      data = bracket_df,
      aes(x = x_start, xend = x_start, y = bracket_y, yend = bracket_y - tick_drop),
      inherit.aes = FALSE,
      linewidth = 0.9
    ) +
    geom_segment(
      data = bracket_df,
      aes(x = x_end, xend = x_end, y = bracket_y, yend = bracket_y - tick_drop),
      inherit.aes = FALSE,
      linewidth = 0.9
    ) +
    geom_text(
      data = bracket_df,
      aes(x = x_mid, y = label_y, label = label),
      inherit.aes = FALSE,
      size = text_size,
      fontface = "bold"
    )
}


deterministic_extra_colors <- function(labels, chroma = 55, luminance = 70) {
  colors <- vapply(labels, function(lbl) {
    ints <- utf8ToInt(enc2utf8(lbl))
    hue <- sum(ints * seq_along(ints)) %% 360
    grDevices::hcl(h = hue, c = chroma, l = luminance)
  }, character(1))
  stats::setNames(colors, labels)
}
```

# Step 2: Define Helper Functions

```{r}
run_shapiro_by_project <- function(data, value_col) {
  project_list <- split(data, data$accession_number)

  bind_rows(
    lapply(
      names(project_list),
      function(project_id) {
        project_df <- project_list[[project_id]]
        bind_rows(
          lapply(
            split(project_df[[value_col]], project_df$group_simple),
            function(values) {
              if (length(values) >= 3 && length(values) <= 5000) {
                test <- shapiro.test(values)
                tibble(
                  n = length(values),
                  statistic = unname(test$statistic),
                  p_value = unname(test$p.value),
                  method = unname(test$method)
                )
              } else {
                tibble(
                  n = length(values),
                  statistic = NA_real_,
                  p_value = NA_real_,
                  method = "Shapiro-Wilk not run; sample size outside 3-5000"
                )
              }
            }
          ),
          .id = "group_simple"
        ) %>%
          mutate(
            accession_number = project_id,
            metric = value_col
          ) %>%
          relocate(accession_number, metric, .before = group_simple)
      }
    )
  )
}

run_mann_whitney_by_project <- function(data, value_col) {
  project_list <- split(data, data$accession_number)

  bind_rows(
    lapply(
      names(project_list),
      function(project_id) {
        project_df <- project_list[[project_id]]

        control_n <- sum(project_df$group_simple == "control", na.rm = TRUE)
        psnp_n <- sum(project_df$group_simple == "psnp", na.rm = TRUE)

        if (control_n > 0 && psnp_n > 0) {
          test <- wilcox.test(
            reformulate("group_simple", response = value_col),
            data = project_df,
            exact = FALSE
          )

          tibble(
            accession_number = project_id,
            metric = value_col,
            n_control = control_n,
            n_psnp = psnp_n,
            statistic = unname(test$statistic),
            p_value = unname(test$p.value),
            method = test$method
          )
        } else {
          tibble(
            accession_number = project_id,
            metric = value_col,
            n_control = control_n,
            n_psnp = psnp_n,
            statistic = NA_real_,
            p_value = NA_real_,
            method = "Mann-Whitney not run; one condition missing"
          )
        }
      }
    )
  )
}

extract_taxonomy_rank <- function(x, rank_letter) {
  pattern <- paste0("(?:^|;)", rank_letter, "_{1,2}([^;]+)")
  value <- stringr::str_match(x, pattern)[, 2]
  value <- ifelse(is.na(value), "", value)
  trimws(gsub("_", " ", value))
}

build_parent_rank_label <- function(family, order, class_name, phylum) {
  invalid_or_generic <- function(v) {
    v == "" | is.na(v) | grepl("^[a-zA-Z]__$", v) |
      tolower(v) %in% c("unassigned", "unclassified", "unknown", "uncultured", "incertae sedis")
  }

  family_invalid <- invalid_or_generic(family)
  order_invalid <- invalid_or_generic(order)
  class_invalid <- invalid_or_generic(class_name)
  phylum_invalid <- invalid_or_generic(phylum)

  parent_taxon <- ifelse(
    !family_invalid, family,
    ifelse(
      !order_invalid, order,
      ifelse(!class_invalid, class_name, ifelse(!phylum_invalid, phylum, NA_character_))
    )
  )

  parent_rank <- ifelse(
    !family_invalid, "family",
    ifelse(
      !order_invalid, "order",
      ifelse(!class_invalid, "class", ifelse(!phylum_invalid, "phylum", NA_character_))
    )
  )

  ifelse(
    !is.na(parent_taxon) & !is.na(parent_rank),
    paste0(parent_rank, ": ", parent_taxon),
    NA_character_
  )
}

clean_taxon_label_followup <- function(x) {
  genus <- extract_taxonomy_rank(x, "g")
  family <- extract_taxonomy_rank(x, "f")
  order <- extract_taxonomy_rank(x, "o")
  class_name <- extract_taxonomy_rank(x, "c")
  phylum <- extract_taxonomy_rank(x, "p")

  genus_invalid <- genus == "" | is.na(genus) | grepl("^[a-z]__$", genus)
  parent_label <- build_parent_rank_label(family, order, class_name, phylum)
  genus_lower <- tolower(genus)

  label <- ifelse(
    !genus_invalid & !genus_lower %in% c("uncultured", "unassigned", "unclassified", "unknown", "incertae sedis"),
    genus,
    ifelse(
      genus_lower == "uncultured",
      ifelse(!is.na(parent_label), paste0("Uncultured (", parent_label, ")"), "Uncultured (taxonomy unavailable)"),
      ifelse(
        genus_lower %in% c("unassigned", "unclassified", "unknown", "incertae sedis"),
        ifelse(!is.na(parent_label), paste0(tools::toTitleCase(genus), " (", parent_label, ")"), tools::toTitleCase(genus)),
        ifelse(!is.na(parent_label), paste0("Unassigned (", parent_label, ")"), "Unassigned (taxonomy unavailable)")
      )
    )
  )

  label <- ifelse(label == "" | is.na(label), "Unclassified", label)
  label
}

build_shared_genus_palette_followup <- function(genus_names) {
  genus_palette <- c(
    "Lactobacillus" = "#8ECFC9",
    "Muribaculaceae" = "#F3C78D",
    "Streptococcus" = "#A8B6D9",
    "Turicibacter" = "#E8A0A6",
    "Faecalibaculum" = "#A9CFA5",
    "Odoribacter" = "#DCC7A1",
    "Bifidobacterium" = "#C8B6D8",
    "Rikenellaceae RC9 gut group" = "#E7B8D6",
    "Alistipes" = "#90B4C6",
    "Helicobacter" = "#D8A48F",
    "Akkermansia" = "#B7A0D8",
    "Bacteroides" = "#8FBE6D",
    "Clostridia UCG-014" = "#B48A8A",
    "Clostridium sensu stricto 1" = "#8D6CCF",
    "Lachnospiraceae NK4A136 group" = "#5B84B1",
    "Muribaculum" = "#9E9D68",
    "Candidatus Saccharimonas" = "#6F8FB4",
    "Desulfovibrio" = "#A77C95",
    "Dubosiella" = "#8AA24A",
    "Enterobacter" = "#E76F51",
    "Blautia" = "#1F78B4",
    "Klebsiella" = "#F39C12",
    "Parasutterella" = "#2F9C95",
    "Candidatus Arthromitus" = "#8E44AD",
    "Incertae Sedis" = "#5FAE8D",
    "Incertae Sedis (Ruminococcaceae)" = "#5FAE8D",
    "RF39" = "#4A90D9",
    "Rikenella" = "#A26CC5",
    "Prevotellaceae UCG-001" = "#2E8B57",
    "[Clostridium] innocuum group" = "#CC8456",
    "Lachnospiraceae UCG-001" = "#5B89C9",
    "[Eubacterium] ventriosum group" = "#C46A93",
    "Uncultured" = "#7BAFD4",
    "Uncultured (Lachnospiraceae)" = "#9FB6C8",
    "Uncultured (Ruminococcaceae)" = "#AFC1D8",
    "Uncultured (Oscillospiraceae)" = "#8AAFD0",
    "Unassigned" = "#C892A6",
    "Unassigned (taxonomy unavailable)" = "#C892A6",
    "Unassigned (Lachnospiraceae)" = "#E0B180",
    "Unclassified" = "#C88EBE",
    "Unclassified (Lachnospiraceae)" = "#D5B3C6",
    "Unclassified (Clostridia)" = "#CFA4BF",
    "Unclassified (Erysipelatoclostridiaceae)" = "#DAB7C5"
  )

  missing_genera <- setdiff(genus_names, names(genus_palette))

  if (length(missing_genera) > 0) {
    extra_cols <- deterministic_extra_colors(missing_genera)
    genus_palette <- c(genus_palette, extra_cols)
  }

  genus_palette
}
```

# Step 3: Read the Current Group 2 Files

```{r}
output_dir <- "/Users/simarjawanda/Library/CloudStorage/OneDrive-UBC/R Analysis/Group 2 Exploratory Outputs"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

alpha_meta_df <- read.csv(
  "/Users/simarjawanda/Library/CloudStorage/OneDrive-UBC/R Analysis/Group 2 Outputs/group2_alpha_diversity_with_metadata.csv",
  stringsAsFactors = FALSE
)

analysis_metadata <- read.csv(
  "/Users/simarjawanda/Library/CloudStorage/OneDrive-UBC/R Analysis/Group 2 Outputs/group2_analysis_metadata.csv",
  stringsAsFactors = FALSE
)

norm_abundance_df <- read.csv(
  "/Users/simarjawanda/Library/CloudStorage/OneDrive-UBC/R Analysis/Group 2 Outputs/group2_abundance_normalized.csv",
  check.names = FALSE,
  stringsAsFactors = FALSE
)

group2_top50_df <- read.csv(
  "/Users/simarjawanda/Library/CloudStorage/OneDrive-UBC/R Analysis/Group 2 Outputs/pca_top50/group2_genus_filtered_top50_for_downstream.csv",
  check.names = FALSE,
  stringsAsFactors = FALSE
)

selected_genus_candidates <- read.csv(
  "/Users/simarjawanda/Library/CloudStorage/OneDrive-UBC/R Analysis/Group 2 Outputs/group2_selected_genus_candidates_top50.csv",
  check.names = FALSE,
  stringsAsFactors = FALSE
)

dim(alpha_meta_df)
dim(group2_top50_df)
selected_genus_candidates
```

# Step 4: Run Per-Project Alpha Exploratory Tests

```{r}
alpha_shapiro_by_project <- bind_rows(
  run_shapiro_by_project(alpha_meta_df, "Shannon"),
  run_shapiro_by_project(alpha_meta_df, "Simpson")
)

alpha_mann_whitney_by_project <- bind_rows(
  run_mann_whitney_by_project(alpha_meta_df, "Shannon"),
  run_mann_whitney_by_project(alpha_meta_df, "Simpson")
)

alpha_shapiro_by_project
alpha_mann_whitney_by_project
```

# Step 5: Make Per-Project Alpha-Diversity Boxplots

```{r}
alpha_plot_data_by_project <- alpha_meta_df %>%
  mutate(
    Condition = ifelse(group_simple == "psnp", "PS-NP", "Control"),
    Condition = factor(Condition, levels = c("Control", "PS-NP")),
    accession_number = factor(accession_number, levels = c("PRJNA944557", "PRJNA1357312"))
  )

shannon_annotation_by_project <- alpha_plot_data_by_project %>%
  distinct(accession_number) %>%
  left_join(
    alpha_mann_whitney_by_project %>%
      filter(metric == "Shannon") %>%
      mutate(label = ifelse(p_value < 0.05, "*", "")) %>%
      select(accession_number, label),
    by = "accession_number"
  )

simpson_annotation_by_project <- alpha_plot_data_by_project %>%
  distinct(accession_number) %>%
  left_join(
    alpha_mann_whitney_by_project %>%
      filter(metric == "Simpson") %>%
      mutate(label = ifelse(p_value < 0.05, "*", "")) %>%
      select(accession_number, label),
    by = "accession_number"
  )

alpha_plot_data_by_project <- alpha_plot_data_by_project %>%
  mutate(
    accession_panel = factor(
      as.character(accession_number),
      levels = c("PRJNA944557", "PRJNA1357312")
    )
  )

shannon_annotation_by_project <- shannon_annotation_by_project %>%
  mutate(
    accession_panel = factor(
      as.character(accession_number),
      levels = c("PRJNA944557", "PRJNA1357312")
    )
  )

simpson_annotation_by_project <- simpson_annotation_by_project %>%
  mutate(
    accession_panel = factor(
      as.character(accession_number),
      levels = c("PRJNA944557", "PRJNA1357312")
    )
  )

shannon_brackets_by_project <- build_two_group_brackets(
  alpha_plot_data_by_project %>%
    left_join(shannon_annotation_by_project %>% select(accession_number, label), by = "accession_number"),
  facet_col = "accession_panel",
  value_col = "Shannon",
  label_col = "label"
)

simpson_brackets_by_project <- build_two_group_brackets(
  alpha_plot_data_by_project %>%
    left_join(simpson_annotation_by_project %>% select(accession_number, label), by = "accession_number"),
  facet_col = "accession_panel",
  value_col = "Simpson",
  label_col = "label"
)

```

```{r group2_exploratory_shannon_boxplot, fig.width=11.5, fig.height=6.8}
group2_exploratory_shannon_boxplot <- ggplot(alpha_plot_data_by_project, aes(x = Condition, y = Shannon, fill = Condition)) +
  stat_boxplot(geom = "errorbar", width = 0.18, linewidth = 0.8, coef = 1.5) +
  geom_boxplot(outlier.shape = NA, width = 0.5, color = "black", linewidth = 0.8, coef = 1.5) +
  geom_jitter(shape = 21, width = 0.08, size = 3.0, color = "black", stroke = 0.45, alpha = 0.8) +
  facet_grid2(cols = vars(accession_panel)) +
  scale_fill_manual(values = c("Control" = "#B7D9F2", "PS-NP" = "#F2B6A7")) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.28))) +
  theme_classic(base_size = 15) +
  labs(
    title = wrap_figure_title("Shannon Alpha Diversity in Group 2 Differentiated by BioProject Accession"),
    x = "Condition",
    y = "Shannon Index"
  ) +
  theme(
    plot.title = element_text(size = 17, hjust = 0.5, face = "plain", lineheight = 1.05, margin = margin(b = 10)),
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 12, color = "black"),
    strip.background = element_blank(),
    strip.text = element_text(size = 13, face = "plain"),
    legend.position = "none"
  )

group2_exploratory_shannon_boxplot <- add_two_group_brackets(group2_exploratory_shannon_boxplot, shannon_brackets_by_project, text_size = 6.8)
group2_exploratory_shannon_boxplot <- finalize_thesis_plot(group2_exploratory_shannon_boxplot, legend = "none") +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 18.5, face = "bold", colour = "black", margin = margin(b = 8)),
    axis.title.y = element_text(size = 23, face = "bold", colour = "black"),
    axis.text.x = element_text(size = 18.5, face = "bold", colour = "black"),
    axis.text.y = element_text(size = 16.8, face = "bold", colour = "black")
  )

group2_exploratory_shannon_boxplot
```

```{r group2_exploratory_simpson_boxplot, fig.width=11.5, fig.height=6.8}
group2_exploratory_simpson_boxplot <- ggplot(alpha_plot_data_by_project, aes(x = Condition, y = Simpson, fill = Condition)) +
  stat_boxplot(geom = "errorbar", width = 0.18, linewidth = 0.8, coef = 1.5) +
  geom_boxplot(outlier.shape = NA, width = 0.5, color = "black", linewidth = 0.8, coef = 1.5) +
  geom_jitter(shape = 21, width = 0.08, size = 3.0, color = "black", stroke = 0.45, alpha = 0.8) +
  facet_grid2(cols = vars(accession_panel)) +
  scale_fill_manual(values = c("Control" = "#B7D9F2", "PS-NP" = "#F2B6A7")) +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.28))) +
  theme_classic(base_size = 15) +
  labs(
    title = wrap_figure_title("Simpson Alpha Diversity in Group 2 Differentiated by BioProject Accession"),
    x = "Condition",
    y = "Simpson Index"
  ) +
  theme(
    plot.title = element_text(size = 17, hjust = 0.5, face = "plain", lineheight = 1.05, margin = margin(b = 10)),
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 12, color = "black"),
    strip.background = element_blank(),
    strip.text = element_text(size = 13, face = "plain"),
    legend.position = "none"
  )

group2_exploratory_simpson_boxplot <- add_two_group_brackets(group2_exploratory_simpson_boxplot, simpson_brackets_by_project, text_size = 6.8)
group2_exploratory_simpson_boxplot <- finalize_thesis_plot(group2_exploratory_simpson_boxplot, legend = "none") +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 18.5, face = "bold", colour = "black", margin = margin(b = 8)),
    axis.title.y = element_text(size = 23, face = "bold", colour = "black"),
    axis.text.x = element_text(size = 18.5, face = "bold", colour = "black"),
    axis.text.y = element_text(size = 16.8, face = "bold", colour = "black")
  )

group2_exploratory_simpson_boxplot
```

# Step 6: Run Per-BioProject Beta Diversity Exploratory Tests

```{r}
colnames(norm_abundance_df)[1] <- "Run"

norm_merged_df <- analysis_metadata %>%
  select(Run, group_simple, accession_number) %>%
  inner_join(norm_abundance_df, by = "Run")

beta_by_project <- bind_rows(
  lapply(
    split(norm_merged_df, norm_merged_df$accession_number),
    function(project_df) {
      abundance_only <- project_df %>%
        select(-Run, -group_simple, -accession_number) %>%
        mutate(across(everything(), as.numeric))

      bray_dist <- vegan::vegdist(abundance_only, method = "bray")
      if (unique(project_df$accession_number) == "PRJNA1357312") {
        set.seed(7)
      } else if (unique(project_df$accession_number) == "PRJNA944557") {
        set.seed(12)
      }

      permanova_df <- as.data.frame(
        vegan::adonis2(bray_dist ~ group_simple, data = project_df, permutations = 999)
      ) %>%
        rownames_to_column("term")

      permanova_df %>%
        mutate(accession_number = unique(project_df$accession_number)) %>%
        relocate(accession_number, .before = term)
    }
  )
)

beta_by_project
```

# Step 7: Run Per-BioProject Genus Mann-Whitney Tests on the Top-50 Table

```{r}
group2_top50_abundance_only <- group2_top50_df %>%
  select(-Run, -group_simple, -psnp_size, -accession_number)

genus_mann_whitney_by_project <- bind_rows(
  lapply(
    split(group2_top50_df, group2_top50_df$accession_number),
    function(project_df) {
      abundance_only <- project_df %>%
        select(-Run, -group_simple, -psnp_size, -accession_number)

      result_df <- bind_rows(
        lapply(
          colnames(abundance_only),
          function(genus_name) {
            test_df <- tibble(
              group_simple = factor(project_df$group_simple, levels = c("control", "psnp")),
              abundance = as.numeric(abundance_only[[genus_name]])
            )

            if (sum(test_df$group_simple == "control") > 0 && sum(test_df$group_simple == "psnp") > 0) {
              test <- wilcox.test(abundance ~ group_simple, data = test_df, exact = FALSE)

              tibble(
                Genus = genus_name,
                p_value = unname(test$p.value)
              )
            } else {
              tibble(
                Genus = genus_name,
                p_value = NA_real_
              )
            }
          }
        )
      ) %>%
        mutate(
          q_value = p.adjust(p_value, method = "BH"),
          accession_number = unique(project_df$accession_number)
        ) %>%
        arrange(p_value)

      result_df %>%
        relocate(accession_number, .before = Genus)
    }
  )
)

genus_mann_whitney_by_project
```

# Step 8: Build the Selected-Genus Plot Tables Differentiated by BioProject Accession

```{r}
selected_genus_boxplot_data <- group2_top50_df %>%
  select(Run, group_simple, accession_number, all_of(selected_genus_candidates$Genus)) %>%
  pivot_longer(
    cols = -c(Run, group_simple, accession_number),
    names_to = "Genus",
    values_to = "RelativeAbundance"
  ) %>%
  mutate(
    RelativeAbundance = as.numeric(RelativeAbundance),
    Condition = ifelse(group_simple == "psnp", "PS-NP", "Control"),
    Condition = factor(Condition, levels = c("Control", "PS-NP")),
    accession_number = factor(accession_number, levels = c("PRJNA944557", "PRJNA1357312")),
    DisplayGenus = clean_taxon_label_followup(Genus)
  ) %>%
  left_join(
    selected_genus_candidates %>%
      select(Genus, DisplayGenus, p_value, q_value, selection_reason, test_type),
    by = c("Genus", "DisplayGenus")
  )

selected_genus_order <- selected_genus_boxplot_data %>%
  group_by(DisplayGenus) %>%
  summarise(TotalAbundance = mean(RelativeAbundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(TotalAbundance)) %>%
  pull(DisplayGenus)

selected_genus_boxplot_data$DisplayGenus <- factor(
  selected_genus_boxplot_data$DisplayGenus,
  levels = selected_genus_order
)

selected_genus_barplot_data <- selected_genus_boxplot_data %>%
  group_by(accession_number, Condition, DisplayGenus) %>%
  summarise(
    MeanRelativeAbundance = mean(RelativeAbundance, na.rm = TRUE),
    p_value = dplyr::first(p_value),
    q_value = dplyr::first(q_value),
    selection_reason = dplyr::first(selection_reason),
    test_type = dplyr::first(test_type),
    .groups = "drop"
  )

selected_genus_stack_order <- selected_genus_barplot_data %>%
  group_by(DisplayGenus) %>%
  summarise(TotalAbundance = sum(MeanRelativeAbundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(TotalAbundance)) %>%
  pull(DisplayGenus)

selected_genus_barplot_data$DisplayGenus <- factor(
  selected_genus_barplot_data$DisplayGenus,
  levels = selected_genus_stack_order
)

selected_genus_palette <- build_shared_genus_palette_followup(levels(selected_genus_barplot_data$DisplayGenus))

selected_genus_annotation <- selected_genus_boxplot_data %>%
  group_by(accession_number, DisplayGenus) %>%
  summarise(
    p_value = dplyr::first(p_value),
    q_value = dplyr::first(q_value),
    label = dplyr::case_when(
      q_value < 0.05 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ ""
    ),
    .groups = "drop"
  )

selected_genus_palette <- build_shared_genus_palette_followup(levels(selected_genus_barplot_data$DisplayGenus))
```

# Step 9: Make the Mean-Abundance Barplot Differentiated by BioProject Accession

```{r group2_selected_genus_barplot_by_project, fig.width=13.5, fig.height=8.4}
selected_genus_barplot_by_project <- ggplot(
  selected_genus_barplot_data,
  aes(x = Condition, y = MeanRelativeAbundance, fill = DisplayGenus)
) +
  geom_col(
    position = position_stack(reverse = TRUE),
    width = 0.55,
    color = "white",
    linewidth = 0.08
  ) +
  facet_grid2(cols = vars(accession_number)) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 0.1),
    breaks = scales::breaks_pretty(n = 8),
    expand = expansion(mult = c(0, 0.18))
  ) +
  scale_fill_manual(
    values = selected_genus_palette,
    breaks = rev(levels(selected_genus_barplot_data$DisplayGenus))
  ) +
  labs(
    title = wrap_figure_title("Mean Relative Abundance of Candidate Genera Differing Between Control and PS-NP Samples in Group 2 Differentiated by BioProject Accession", width = 62),
    x = "Condition",
    y = "Mean Relative Abundance (%)",
    fill = "Genus"
  ) +
  theme_classic(base_size = 15) +
  theme(
    plot.title = element_text(size = 18, hjust = 0.5, face = "plain", lineheight = 1.05, margin = margin(b = 12)),
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 12, color = "black"),
    strip.background = element_blank(),
    strip.text = element_text(size = 13, face = "plain"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 9.5, face = "italic"),
    legend.position = "right",
    legend.key.height = unit(0.3, "cm"),
    legend.spacing.y = unit(0.06, "cm"),
    plot.title.position = "plot"
  )

selected_genus_barplot_by_project <- finalize_thesis_plot(selected_genus_barplot_by_project, legend = "italic") +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 15.5, face = "bold", colour = "black", margin = margin(b = 6)),
    axis.title.y = element_text(size = 18.5, face = "bold", colour = "black"),
    axis.text.x = element_text(size = 15.5, face = "bold", colour = "black")
  )

selected_genus_barplot_by_project
```

# Step 10: Save the Exploratory Outputs

```{r}
write.csv(
  alpha_shapiro_by_project,
  file.path(output_dir, "group2_exploratory_shapiro_by_project.csv"),
  row.names = FALSE
)

write.csv(
  alpha_mann_whitney_by_project,
  file.path(output_dir, "group2_exploratory_alpha_mann_whitney_by_project.csv"),
  row.names = FALSE
)

write.csv(
  alpha_plot_data_by_project,
  file.path(output_dir, "group2_exploratory_alpha_boxplot_data_by_project.csv"),
  row.names = FALSE
)

write.csv(
  beta_by_project,
  file.path(output_dir, "group2_exploratory_permanova_by_project.csv"),
  row.names = FALSE
)

write.csv(
  genus_mann_whitney_by_project,
  file.path(output_dir, "group2_exploratory_genus_mann_whitney_by_project_top50.csv"),
  row.names = FALSE
)

write.csv(
  selected_genus_boxplot_data,
  file.path(output_dir, "group2_exploratory_selected_genus_boxplot_data_by_project.csv"),
  row.names = FALSE
)

write.csv(
  selected_genus_barplot_data,
  file.path(output_dir, "group2_exploratory_selected_genus_barplot_data_by_project.csv"),
  row.names = FALSE
)

ggsave(
  file.path(output_dir, "group2_exploratory_shannon_boxplot_by_project.png"),
  plot = group2_exploratory_shannon_boxplot,
  width = 11.5,
  height = 6.8,
  dpi = 600
)

ggsave(
  file.path(output_dir, "group2_exploratory_simpson_boxplot_by_project.png"),
  plot = group2_exploratory_simpson_boxplot,
  width = 11.5,
  height = 6.8,
  dpi = 600
)

ggsave(
  file.path(output_dir, "group2_exploratory_selected_genus_barplot_by_project.png"),
  plot = selected_genus_barplot_by_project,
  width = 13.5,
  height = 8.4,
  dpi = 600
)
```

This exploratory notebook keeps the main pooled Group 2 figures clean, while showing how the key top-50 candidate genera and core alpha/beta summaries behave when differentiated by BioProject accession.
