---
title: "Group 4 LEfSe Exploratory Analysis (LDA = 1)"
author: "Simar Jawanda"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  dev = "png",
  fig.ext = "png",
  dpi = 300
)
```

# Step 1: Load Packages

```{r}
Sys.setenv(
  PATH = paste("/opt/homebrew/bin", Sys.getenv("PATH"), sep = ":"),
  FC = "/opt/homebrew/bin/gfortran",
  F77 = "/opt/homebrew/bin/gfortran",
  F90 = "/opt/homebrew/bin/gfortran"
)

library(lefser)
library(SummarizedExperiment)
library(dplyr)
```

# Step 2: Read the Saved Top-50 Filtered Table

```{r}
genus_merged_df <- read.csv(
  "/Users/simarjawanda/Library/CloudStorage/OneDrive-UBC/R Analysis/Group 4 Outputs/pca_top50/group4_genus_filtered_top50_for_downstream.csv",
  check.names = FALSE,
  stringsAsFactors = FALSE
)

dim(genus_merged_df)
```

# Step 3: Prepare the Top-50 Genus Matrix

```{r}
genus_abundance_only <- genus_merged_df %>%
  select(-Run, -group_simple, -psnp_size, -accession_number)

dim(genus_abundance_only)
```

# Step 4: Prepare LEfSe Input

```{r}
lefse_meta <- genus_merged_df %>%
  select(Run, group_simple) %>%
  mutate(
    group_simple = factor(trimws(tolower(group_simple)), levels = c("control", "psnp"))
  ) %>%
  filter(group_simple %in% c("control", "psnp"))

lefse_genus <- genus_abundance_only[
  match(lefse_meta$Run, genus_merged_df$Run),
  ,
  drop = FALSE
]

lefse_matrix <- t(as.matrix(lefse_genus))
mode(lefse_matrix) <- "numeric"

lefse_input <- SummarizedExperiment(
  assays = list(relative_abundance = lefse_matrix),
  colData = S4Vectors::DataFrame(lefse_meta)
)

dim(lefse_matrix)
table(lefse_meta$group_simple)
```

# Step 5: Run LEfSe with LDA = 1

```{r}
group4_lefse_lda1 <- lefser(
  lefse_input,
  classCol = "group_simple",
  assay = "relative_abundance",
  lda.threshold = 1,
  method = "none",
  trim.names = FALSE
)

group4_lefse_lda1 <- as.data.frame(group4_lefse_lda1)

if (nrow(group4_lefse_lda1) > 0) {
  colnames(group4_lefse_lda1) <- c("Genus", "LDA_score")
  group4_lefse_lda1 <- group4_lefse_lda1 %>%
    arrange(desc(LDA_score))
} else {
  group4_lefse_lda1 <- data.frame(
    Genus = character(),
    LDA_score = numeric()
  )
}

group4_lefse_lda1
```

# Step 6: Read the Current Top-50 Genus Results

```{r}
current_results_df <- read.csv(
  "/Users/simarjawanda/Library/CloudStorage/OneDrive-UBC/R Analysis/Group 4 Outputs/pca_top50/group4_genus_level_mann_whitney_fdr_results_top50.csv",
  stringsAsFactors = FALSE
)

dim(current_results_df)
```

# Step 7: Compare LEfSe with the Current Top-50 Mann-Whitney Results

```{r}
group4_lefse_lda1_compare <- current_results_df %>%
  mutate(
    Genus = as.character(Genus),
    Found_by_LEfSe_LDA1 = Genus %in% group4_lefse_lda1$Genus,
    LEfSe_LDA1_score = group4_lefse_lda1$LDA_score[match(Genus, group4_lefse_lda1$Genus)],
    FDR_significant = q_value < 0.05
  ) %>%
  arrange(p_value)

head(group4_lefse_lda1_compare, 20)
```

# Step 8: Save Top-50 LDA = 1 LEfSe Outputs

```{r}
lefse_output_dir <- "/Users/simarjawanda/Library/CloudStorage/OneDrive-UBC/R Analysis/LEfSe LDA 1 Top50/Group_4"
dir.create(lefse_output_dir, recursive = TRUE, showWarnings = FALSE)

write.csv(
  group4_lefse_lda1,
  file.path(lefse_output_dir, "group4_lefse_results_LDA1_top50.csv"),
  row.names = FALSE
)

write.csv(
  group4_lefse_lda1_compare,
  file.path(lefse_output_dir, "group4_lefse_vs_mann_whitney_top50_comparison_LDA1.csv"),
  row.names = FALSE
)

list.files(lefse_output_dir, pattern = "LDA1", full.names = FALSE)
```

This exploratory notebook checks whether lowering the LEfSe LDA threshold from 2 to 1 reveals any biomarkers in the finalized top-50 workflow and saves the labeled outputs separately from the main analysis.
